---
title: "Leukemia Target Discovery and Validation Pipeline"
author: "Qinuo Shi & Antigravity"
date: "2026-06-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

# Complete Bioinformatics Pipeline

This document integrates the complete target discovery, safety filtering, clinical prognosis association, pathway enrichment, and cross-dataset validation pipeline.

## Section 1: Download TCGA-LAML Dataset

Downloads and formats the TCGA Acute Myeloid Leukemia dataset from GDC, applying the getBarcodeInfo and GDCquery_clinic namespace patches to bypass schema changes.

```{r}
## ----setup, include=FALSE-----------------------------------------------------
knitr::opts_chunk$set(echo = TRUE)


## -----------------------------------------------------------------------------
# scripts/01_download_tcga_laml.R

# ==============================================================================
# SECTION 0: Setup and Package Loading | 第 0 部分：准备与加载包
# ------------------------------------------------------------------------------
# [EN] Check for required BioConductor and CRAN packages, install if missing, and load libraries.
# [ZH] 检查所需的 BioConductor 和 CRAN 依赖包，若本地未安装则进行自动安装，并加载相应的 R 库。
# ==============================================================================

if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

required_bioc <- c("TCGAbiolinks", "SummarizedExperiment")
required_cran <- c("data.table", "dplyr", "stringr")

for (pkg in required_bioc) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    BiocManager::install(pkg)
  }
}

for (pkg in required_cran) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    install.packages(pkg)
  }
}

library(TCGAbiolinks)
library(SummarizedExperiment)
library(data.table)
library(dplyr)
library(stringr)


# ==============================================================================
# SECTION 1: In-Memory Namespace Patches for TCGAbiolinks | 第 1 部分：TCGAbiolinks 内存命名空间补丁
# ------------------------------------------------------------------------------
# [EN] Dynamic patch to override internal getBarcodeInfo & exported GDCquery_clinic
#      functions to bypass the GDC API schema changes causing "disease_response doesn't exist" errors.
# [ZH] 动态修补包内部 getBarcodeInfo 与 GDCquery_clinic 函数，绕过因 GDC API 字段变更（缺少 disease_response 列）导致的报错。
# ==============================================================================

ns <- asNamespace("TCGAbiolinks")
pkg_env <- as.environment("package:TCGAbiolinks")

# 1.1 Patch getBarcodeInfo (used internally by GDCprepare)
if (exists("getBarcodeInfo", envir = ns)) {
  unlockBinding("getBarcodeInfo", ns)
  
  patched_getBarcodeInfo <- function (barcode) 
  {
      baseURL <- "https://api.gdc.cancer.gov/cases/?"
      options.pretty <- "pretty=true"
      options.expand <- "expand=samples,follow_ups,project,diagnoses,diagnoses.treatments,annotations,family_histories,demographic,exposures"
      option.size <- paste0("size=", length(barcode))
      options.filter <- paste0("filters=", URLencode("{\"op\":\"or\",\"content\":[{\"op\":\"in\",\"content\":{\"field\":\"cases.submitter_id\",\"value\":["), 
          paste0("\"", paste(barcode, collapse = "\",\"")), URLencode("\"]}},"), 
          URLencode("{\"op\":\"in\",\"content\":{\"field\":\"submitter_sample_ids\",\"value\":["), 
          paste0("\"", paste(barcode, collapse = "\",\"")), URLencode("\"]}},"), 
          URLencode("{\"op\":\"in\",\"content\":{\"field\":\"submitter_aliquot_ids\",\"value\":["), 
          paste0("\"", paste(barcode, collapse = "\",\"")), URLencode("\"]}}"), 
          URLencode("]}"))
      url <- paste0(baseURL, paste(options.pretty, options.expand, 
          option.size, options.filter, sep = "&"))
      
      getURL <- TCGAbiolinks:::getURL
      json <- tryCatch(getURL(url, jsonlite::fromJSON, httr::timeout(600), simplifyDataFrame = TRUE), 
          error = function(e) {
              message(paste("Error: ", e, sep = " "))
              message("We will retry to access GDC again! URL:")
              jsonlite::fromJSON(httr::content(getURL(url, httr::GET, httr::timeout(600)), 
                  as = "text", encoding = "UTF-8"), simplifyDataFrame = TRUE)
          })
      results <- json$data$hits
      if (length(results) == 0) {
          return(data.frame(barcode, stringsAsFactors = FALSE))
      }
      submitter_id <- results$submitter_id
      submitter_aliquot_ids <- results$submitter_aliquot_ids
      
      df <- data.frame(submitter_id = submitter_id, stringsAsFactors = FALSE)
      
      if (!is.null(results$samples)) {
          samples <- data.table::rbindlist(results$samples, fill = TRUE)
          samples <- samples[match(barcode, samples$submitter_id), ]
          samples$sample_submitter_id <- stringr::str_extract_all(samples$submitter_id, 
              paste(barcode, collapse = "|")) %>% unlist %>% as.character
          tryCatch({
              samples$submitter_id <- stringr::str_extract_all(samples$submitter_id, 
                  paste(c(submitter_id, barcode), collapse = "|"), 
                  simplify = TRUE) %>% as.character
          }, error = function(e) {
              samples$submitter_id <- submitter_id
          })
          df <- samples[!is.na(samples$submitter_id), ]
          suppressWarnings({
              df[, c("updated_datetime", "created_datetime")] <- NULL
          })
      }
      if (!is.null(results$diagnoses)) {
          diagnoses <- data.table::rbindlist(lapply(results$diagnoses, function(x) if (is.null(x)) 
              data.frame(NA)
          else x), fill = TRUE)
          cols_to_remove <- intersect(c("updated_datetime", "created_datetime", "state", "days_to_last_follow_up"), colnames(diagnoses))
          for (col in cols_to_remove) {
              diagnoses[[col]] <- NULL
          }
          if (any(grepl("submitter_id", colnames(diagnoses)))) {
              diagnoses$submitter_id <- gsub("-diagnosis|_diagnosis.*|-DIAG|diag-", 
                  "", diagnoses$submitter_id)
          }
          else {
              diagnoses$submitter_id <- submitter_id
          }
          if (!any(df$submitter_id %in% diagnoses$submitter_id)) {
              diagnoses$submitter_id <- NULL
              df <- dplyr::bind_cols(df %>% as.data.frame, diagnoses %>% 
                  as.data.frame)
          }
          else {
              df <- dplyr::left_join(df, diagnoses, by = "submitter_id")
          }
      }
      if (!is.null(results$exposures)) {
          exposures <- data.table::rbindlist(lapply(results$exposures, function(x) if (is.null(x)) 
              data.frame(NA)
          else x), fill = TRUE)
          cols_to_remove <- intersect(c("updated_datetime", "created_datetime", "state"), colnames(exposures))
          for (col in cols_to_remove) {
              exposures[[col]] <- NULL
          }
          if (any(grepl("submitter_id", colnames(exposures)))) {
              exposures$submitter_id <- gsub("-exposure|_exposure.*|-EXP", 
                  "", exposures$submitter_id)
          }
          else {
              exposures$submitter_id <- submitter_id
          }
          if (!any(df$submitter_id %in% exposures$submitter_id)) {
              exposures$submitter_id <- NULL
              df <- dplyr::bind_cols(df, exposures)
          }
          else {
              df <- dplyr::left_join(df, exposures, by = "submitter_id")
          }
      }
      if (!is.null(results$follow_ups)) {
          follow_ups <- data.table::rbindlist(lapply(results$follow_ups, function(x) if (is.null(x)) 
              data.frame(NA)
          else x), fill = TRUE)
          cols_to_remove <- intersect(c("updated_datetime", "created_datetime", "state"), colnames(follow_ups))
          for (col in cols_to_remove) {
              follow_ups[[col]] <- NULL
          }
          if (any(grepl("submitter_id", colnames(follow_ups)))) {
              follow_ups$submitter_id <- gsub("_follow_up.*", "", 
                  follow_ups$submitter_id)
              
              # Safe selection & rename
              select_cols <- c("submitter_id", "days_to_follow_up")
              if ("disease_response" %in% colnames(follow_ups)) {
                  select_cols <- c(select_cols, "disease_response")
              }
              
              if (all(c("submitter_id", "days_to_follow_up") %in% colnames(follow_ups))) {
                  follow_ups_last <- follow_ups %>% dplyr::select(dplyr::all_of(select_cols)) %>% 
                      dplyr::filter(!is.na(submitter_id), !is.na(days_to_follow_up))
                  
                  if (nrow(follow_ups_last) > 0) {
                      follow_ups_last <- follow_ups_last %>% dplyr::group_by(submitter_id) %>% 
                          dplyr::filter(dplyr::row_number() == which.max(days_to_follow_up)) %>% 
                          dplyr::ungroup()
                      
                      if ("disease_response" %in% colnames(follow_ups_last)) {
                          follow_ups_last <- follow_ups_last %>% 
                              dplyr::rename_at(dplyr::vars(disease_response), .funs = function(x) paste0("follow_ups_", x))
                      }
                      
                      follow_ups_last <- follow_ups_last %>% 
                          dplyr::rename(days_to_last_follow_up = days_to_follow_up)
                      df <- dplyr::left_join(df, follow_ups_last, by = "submitter_id")
                  }
              }
          }
      }
      if (!is.null(results$demographic)) {
          demographic <- results$demographic
          cols_to_remove <- intersect(c("updated_datetime", "created_datetime", "state"), colnames(demographic))
          for (col in cols_to_remove) {
              demographic[[col]] <- NULL
          }
          if (any(grepl("submitter_id", colnames(demographic)))) {
              demographic$submitter_id <- gsub("-demographic|_demographic.*|-DEMO|demo-", 
                  "", results$demographic$submitter_id)
          }
          else {
              demographic$submitter_id <- submitter_id
          }
          if (!any(df$submitter_id %in% demographic$submitter_id)) {
              demographic$submitter_id <- NULL
              if (nrow(demographic) < nrow(df)) {
                  demographic <- plyr::ldply(1:length(results$submitter_sample_ids), 
                    .fun = function(x) {
                      demographic[x, ] %>% as.data.frame() %>% 
                        dplyr::slice(rep(dplyr::row_number(), sum(results$submitter_sample_ids[[x]] %in% 
                          barcode)))
                    })
              }
              df <- dplyr::bind_cols(df %>% as.data.frame, demographic)
          }
          else {
              df <- dplyr::left_join(df, demographic, by = "submitter_id")
          }
      }
      df$bcr_patient_barcode <- df$submitter_id %>% as.character()
      projects.info <- results$project
      projects.info <- results$project[, grep("state", colnames(projects.info), 
          invert = TRUE)]
      if (any(submitter_id %in% df$submitter_id)) {
          projects.info <- cbind(submitter_id = submitter_id, projects.info)
          suppressWarnings({
              df <- dplyr::left_join(df, projects.info, by = "submitter_id")
          })
      }
      else {
          if (nrow(projects.info) < nrow(df)) {
              projects.info <- plyr::ldply(1:length(results$submitter_sample_ids), 
                  .fun = function(x) {
                    projects.info[x, ] %>% as.data.frame() %>% 
                      dplyr::slice(rep(dplyr::row_number(), sum(results$submitter_sample_ids[[x]] %in% 
                        barcode)))
                  })
          }
          df <- dplyr::bind_cols(df, projects.info)
      }
      if (any(substr(barcode, 1, stringr::str_length(df$submitter_id)) %in% 
          df$submitter_id)) {
          df <- df[match(substr(barcode, 1, stringr::str_length(df$sample_submitter_id)), 
              df$sample_submitter_id), ]
          df <- df[!is.na(df$submitter_id), ]
      }
      else {
          idx <- sapply(substr(barcode, 1, stringr::str_length(df$submitter_aliquot_ids) %>% 
              max), FUN = function(x) {
              grep(x, df$submitter_aliquot_ids)
          })
          df <- df[idx, ]
      }
      df <- df %>% as.data.frame() %>% dplyr::select(which(colSums(is.na(df)) < 
          nrow(df)))
      return(df)
  }
  
  assign("getBarcodeInfo", patched_getBarcodeInfo, envir = ns)
  lockBinding("getBarcodeInfo", ns)
}

# 1.2 Patch GDCquery_clinic (used directly to fetch clinical data)
if (exists("GDCquery_clinic", envir = ns)) {
  fn_lines <- deparse(TCGAbiolinks::GDCquery_clinic)
  start_idx <- grep("follow_ups_last <- follow_ups %>% dplyr::select", fn_lines, fixed = TRUE)
  end_idx <- grep("df <- dplyr::full_join(df, follow_ups_last", fn_lines, fixed = TRUE) - 1
  
  if (length(start_idx) == 1 && length(end_idx) == 1) {
    patched_code <- c(
      "                select_cols <- c('submitter_id', 'days_to_follow_up')",
      "                if ('disease_response' %in% colnames(follow_ups)) {",
      "                    select_cols <- c(select_cols, 'disease_response')",
      "                }",
      "                follow_ups_last <- follow_ups %>% dplyr::select(dplyr::all_of(select_cols)) %>% ",
      "                  dplyr::filter(!is.na(submitter_id), !is.na(days_to_follow_up))",
      "                if (nrow(follow_ups_last) > 0) {",
      "                    follow_ups_last <- follow_ups_last %>% dplyr::group_by(submitter_id) %>% ",
      "                      dplyr::filter(dplyr::row_number() == which.max(days_to_follow_up)) %>% ",
      "                      dplyr::ungroup()",
      "                    if ('disease_response' %in% colnames(follow_ups_last)) {",
      "                        follow_ups_last <- follow_ups_last %>% dplyr::rename_at(dplyr::vars(disease_response), .funs = function(x) paste0('follow_ups_', x))",
      "                    }",
      "                    follow_ups_last <- follow_ups_last %>% dplyr::rename(days_to_last_follow_up = days_to_follow_up)",
      "                } else {",
      "                    follow_ups_last <- data.frame(submitter_id = character(), days_to_last_follow_up = numeric(), stringsAsFactors = FALSE)",
      "                }"
    )
    
    patched_fn_lines <- c(fn_lines[1:(start_idx-1)], patched_code, fn_lines[(end_idx+1):length(fn_lines)])
    patched_GDCquery_clinic <- eval(parse(text = paste(patched_fn_lines, collapse = "\n")))
    
    # Set environment to package namespace
    environment(patched_GDCquery_clinic) <- ns
    
    # Update Namespace
    unlockBinding("GDCquery_clinic", ns)
    assign("GDCquery_clinic", patched_GDCquery_clinic, envir = ns)
    lockBinding("GDCquery_clinic", ns)
    
    # Update Package Environment (Search Path)
    if (exists("GDCquery_clinic", envir = pkg_env)) {
      unlockBinding("GDCquery_clinic", pkg_env)
      assign("GDCquery_clinic", patched_GDCquery_clinic, envir = pkg_env)
      lockBinding("GDCquery_clinic", pkg_env)
    }
  }
}


# ==============================================================================
# SECTION 2: Create Output Folders | 第 2 部分：创建输出文件夹
# ------------------------------------------------------------------------------
# [EN] Sets up local directories for saving raw downloads and cleaned results.
# [ZH] 创建本地文件夹，分别用于存储下载的原始数据和最终清洗出的结果数据。
# ==============================================================================

dir.create("data_raw/tcga_laml", recursive = TRUE, showWarnings = FALSE)
dir.create("data_clean", recursive = TRUE, showWarnings = FALSE)


# ==============================================================================
# SECTION 3: Query TCGA-LAML Gene Expression Data | 第 3 部分：构建 TCGA-LAML 表达数据查询
# ------------------------------------------------------------------------------
# [EN] Configure query filters on the GDC portal to retrieve transcriptomic RNA-Seq (STAR - Counts) files for TCGA-LAML.
# [ZH] 配置 GDC 数据库查询过滤器，定位 TCGA-LAML 项目的转录组定量数据（STAR - Counts 类型）。
# ==============================================================================

query_expr <- GDCquery(
  project = "TCGA-LAML",
  data.category = "Transcriptome Profiling",
  data.type = "Gene Expression Quantification",
  workflow.type = "STAR - Counts"
)


# ==============================================================================
# SECTION 4: Download TCGA-LAML Raw Files | 第 4 部分：下载 TCGA-LAML 原始数据
# ------------------------------------------------------------------------------
# [EN] Downloads individual sample-level expression TSV files based on the query to local raw folder.
# [ZH] 执行数据下载，将查询到的每个样本的表达定量 TSV 散装文件下载至本地原始数据目录中。
# ==============================================================================

GDCdownload(
  query = query_expr,
  directory = "data_raw/tcga_laml"
)


# ==============================================================================
# SECTION 5: Merge Expression Matrix & Extract Gene Annotations | 第 5 部分：整理表达矩阵与基因注释
# ------------------------------------------------------------------------------
# [EN] Uses GDCprepare() to read and merge downloaded TSVs into a single TPM matrix and extract gene metadata.
# [ZH] 通过 GDCprepare() 批量读取下载的样本 TSV 并自动拼接合并为完整的 TPM 表达量矩阵，同时提取基因元数据。
# ==============================================================================

data_prep <- GDCprepare(
  query = query_expr,
  directory = "data_raw/tcga_laml"
)

# Extract TPM matrix (STAR unstranded TPM)
tcga_tpm <- assay(data_prep, "tpm_unstrand")

# Check TPM dimensions and columns
cat("Dimension of TCGA-LAML TPM matrix:\n")
print(dim(tcga_tpm))

cat("Preview of TPM matrix:\n")
print(tcga_tpm[1:5, 1:min(5, ncol(tcga_tpm))])

# Save raw TPM expression matrix
write.csv(
  tcga_tpm,
  file = "data_raw/tcga_laml/tcga_laml_tpm_raw.csv"
)

saveRDS(
  tcga_tpm,
  file = "data_raw/tcga_laml/tcga_laml_tpm_raw.rds"
)

# Save gene annotation information
gene_info <- as.data.frame(rowRanges(data_prep))
# Clean gene info format slightly (matching original)
gene_info <- data.frame(
  gene_id = gene_info$gene_id,
  gene_name = gene_info$gene_name,
  gene_type = gene_info$gene_type,
  stringsAsFactors = FALSE
)

write.csv(
  gene_info,
  file = "data_raw/tcga_laml/tcga_laml_gene_info.csv",
  row.names = FALSE
)


# ==============================================================================
# SECTION 6: Download & Format Clinical Data | 第 6 部分：下载与整理临床数据
# ------------------------------------------------------------------------------
# [EN] Downloads patient clinical tables using GDCquery_clinic() and formats list columns for CSV compatibility.
# [ZH] 使用 GDCquery_clinic() 下载完整的患者临床信息，并将嵌套的 List 类型列进行字符化展平，以支持 CSV 导出。
# ==============================================================================

tcga_clinical <- GDCquery_clinic(
  project = "TCGA-LAML",
  type = "clinical"
)

# Convert list columns to character for CSV export (preserves data while allowing CSV writing)
tcga_clinical_csv <- tcga_clinical
list_cols <- sapply(tcga_clinical_csv, is.list)
for (col in names(list_cols)[list_cols]) {
  tcga_clinical_csv[[col]] <- sapply(tcga_clinical_csv[[col]], function(x) paste(x, collapse = ";"))
}

write.csv(
  tcga_clinical_csv,
  file = "data_raw/tcga_laml/tcga_laml_clinical.csv",
  row.names = FALSE
)

saveRDS(
  tcga_clinical,
  file = "data_raw/tcga_laml/tcga_laml_clinical.rds"
)

cat("Clinical data downloaded using TCGAbiolinks.\n")
cat("Dimension of clinical data:\n")
print(dim(tcga_clinical))

cat("Clinical data preview:\n")
print(head(tcga_clinical))


# ==============================================================================
# SECTION 7: Aggregate Duplicates by Unversioned Ensembl ID & Transform to log2(TPM + 1) | 第 7 部分：去版本号合并重复基因与对数转化
# ------------------------------------------------------------------------------
# [EN] Strips version suffix from Ensembl IDs, aggregates duplicate genes (e.g. PAR genes)
#      by mean TPM, and applies log2(TPM + 1) transformation.
# [ZH] 去除 Ensembl ID 的版本号后缀，对重复基因（如 PAR 拟常染色体基因区域）取 TPM 均值合并，并进行 log2(TPM + 1) 转化。
# ==============================================================================

# Convert TCGA matrix to data.table for fast aggregation
tcga_tpm_dt <- as.data.table(tcga_tpm, keep.rownames = "ensembl_id")
tcga_tpm_dt[, ensembl_id_clean := str_remove(ensembl_id, "\\..*$")]

# Retrieve expression columns
tcga_expr_cols <- setdiff(colnames(tcga_tpm_dt), c("ensembl_id", "ensembl_id_clean"))

# Aggregate by clean Ensembl ID (taking the mean of duplicates)
tcga_tpm_by_gene <- tcga_tpm_dt[, lapply(.SD, mean, na.rm = TRUE), by = ensembl_id_clean, .SDcols = tcga_expr_cols]

# Format as data.frame with clean Ensembl IDs as row names
tcga_expr_df <- as.data.frame(tcga_tpm_by_gene)
rownames(tcga_expr_df) <- tcga_expr_df$ensembl_id_clean
tcga_expr_df$ensembl_id_clean <- NULL

tcga_log2_tpm <- log2(tcga_expr_df + 1)

write.csv(
  tcga_log2_tpm,
  file = "data_clean/tcga_laml_expr_log2tpm.csv"
)

saveRDS(
  tcga_log2_tpm,
  file = "data_clean/tcga_laml_expr_log2tpm.rds"
)



# ==============================================================================
# SECTION 8: Save Group and Metadata Mapping | 第 8 部分：保存样本元数据映射
# ------------------------------------------------------------------------------
# [EN] Exports a mapping table of sample barcodes defining group (AML) and source (TCGA) groups.
# [ZH] 导出样本映射元数据表格，指定疾病分组（AML）和样本来源数据库（TCGA）。
# ==============================================================================

tcga_sample_info <- data.frame(
  sample_id = colnames(tcga_tpm),
  source = "TCGA",
  group = "AML"
)

write.csv(
  tcga_sample_info,
  file = "data_clean/tcga_laml_sample_info.csv",
  row.names = FALSE
)

cat("TCGA-LAML download and preprocessing finished.\n")


```

## Section 2: Download GTEx Normal Controls

Downloads GTEx V8 whole blood expression matrix and sample metadata from Google Cloud storage, filtering for normal blood controls.

```{r}
# ==============================================================================
# SECTION 0: Load Required Packages | 第 0 部分：加载所需的 R 包
# ------------------------------------------------------------------------------
# [EN] Check for and install missing CRAN packages, then load library dependencies for GTEx.
# [ZH] 检查并自动安装缺失的 CRAN 包，然后加载处理 GTEx 数据所需的相应 R 库。
# ==============================================================================

required_cran <- c("data.table", "dplyr", "stringr")

for (pkg in required_cran) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    install.packages(pkg)
  }
}

library(data.table)
library(dplyr)
library(stringr)


# ==============================================================================
# SECTION 1: Create GTEx Output Folders | 第 1 部分：创建 GTEx 输出目录
# ------------------------------------------------------------------------------
# [EN] Initialize local directory structures for raw GTEx downloads and clean output matrices.
# [ZH] 创建本地文件夹结构，分别用于存放 GTEx 的原始下载数据和清洗后的最终数据。
# ==============================================================================

dir.create("data_raw/gtex", recursive = TRUE, showWarnings = FALSE)
dir.create("data_clean", recursive = TRUE, showWarnings = FALSE)


# ==============================================================================
# SECTION 2: Define GTEx V8 URLs and File Paths | 第 2 部分：定义 GTEx V8 下载链接与本地路径
# ------------------------------------------------------------------------------
# [EN] Specify URLs and local destinations for GTEx V8 expression data and sample attributes.
# [ZH] 指定 GTEx V8 表达量数据和样本属性数据的 Google Cloud 存储链接及对应的本地保存路径。
# ==============================================================================

gtex_tpm_url <- "https://storage.googleapis.com/adult-gtex/bulk-gex/v8/rna-seq/GTEx_Analysis_2017-06-05_v8_RNASeQCv1.1.9_gene_tpm.gct.gz"

gtex_sample_url <- "https://storage.googleapis.com/adult-gtex/annotations/v8/metadata-files/GTEx_Analysis_v8_Annotations_SampleAttributesDS.txt"

gtex_tpm_file <- "data_raw/gtex/GTEx_gene_tpm.gct.gz"
gtex_sample_file <- "data_raw/gtex/GTEx_sample_attributes.txt"


# ==============================================================================
# SECTION 3: Download GTEx Dataset | 第 3 部分：执行 GTEx 数据下载
# ------------------------------------------------------------------------------
# [EN] Set connection timeout and download GTEx expression and annotation files if missing.
# [ZH] 调整下载超时限制，并自动下载缺损的 GTEx 表达矩阵和样本属性文件。
# ==============================================================================

options(timeout = 100000)

download_if_missing <- function(url, destfile) {
  if (!file.exists(destfile)) {
    cat("Downloading:", destfile, "\n")
    download.file(
      url = url,
      destfile = destfile,
      mode = "wb"
    )
  } else {
    cat("File already exists:", destfile, "\n")
  }
}

download_if_missing(gtex_sample_url, gtex_sample_file)
download_if_missing(gtex_tpm_url, gtex_tpm_file)


# ==============================================================================
# SECTION 4: Load and Inspect Sample Attributes | 第 4 部分：加载并查看样本属性注释
# ------------------------------------------------------------------------------
# [EN] Read sample attributes table with data.table and export all unique tissue names.
# [ZH] 使用 data.table 快速读取样本注释表格，并保存数据集中所有可用的组织类型名称。
# ==============================================================================

gtex_sample_attr <- fread(gtex_sample_file)

cat("GTEx sample annotation dimensions:\n")
print(dim(gtex_sample_attr))

cat("GTEx sample annotation columns:\n")
print(colnames(gtex_sample_attr))

# Save available tissue names for checking
available_tissues <- sort(unique(gtex_sample_attr$SMTSD))

write.csv(
  data.frame(tissue = available_tissues),
  file = "data_raw/gtex/gtex_available_tissues.csv",
  row.names = FALSE
)

cat("Available GTEx tissues were saved to data_raw/gtex/gtex_available_tissues.csv\n")


# ==============================================================================
# SECTION 5: Filter Normal Control Tissues | 第 5 部分：筛选对照组正常组织
# ------------------------------------------------------------------------------
# [EN] Match target tissues (Whole Blood and Bone Marrow) and retrieve corresponding sample IDs.
# [ZH] 匹配目标对照组织（全血与骨髓），并提取其在 GTEx 数据库中对应的所有样本 ID。
# ==============================================================================

target_tissues <- c("Whole Blood", "Bone Marrow")

available_target_tissues <- intersect(target_tissues, available_tissues)
missing_target_tissues <- setdiff(target_tissues, available_tissues)

cat("Target tissues found in GTEx:\n")
print(available_target_tissues)

cat("Target tissues not found in GTEx:\n")
print(missing_target_tissues)

if (length(available_target_tissues) == 0) {
  stop("None of the target tissues were found in GTEx sample annotation.")
}

gtex_target_sample_info <- gtex_sample_attr %>%
  filter(SMTSD %in% available_target_tissues)

target_sample_ids <- gtex_target_sample_info$SAMPID

cat("Number of selected GTEx samples:\n")
print(length(target_sample_ids))

cat("Selected tissue counts:\n")
print(table(gtex_target_sample_info$SMTSD))


# ==============================================================================
# SECTION 6: Read Selected Samples from TPM Matrix | 第 6 部分：筛选并加载目标样本表达矩阵
# ------------------------------------------------------------------------------
# [EN] Scan TPM matrix header to match sample IDs and load only relevant expression columns.
# [ZH] 预读取大表达矩阵头部以匹配样本 ID，仅加载指定正常样本的表达量数据以节省内存。
# ==============================================================================

# GCT files have two metadata rows, so we use skip = 2.
# First, read only the header to avoid loading the full large matrix.
gtex_header <- fread(gtex_tpm_file, skip = 2, nrows = 0)

gct_columns <- colnames(gtex_header)

sample_ids_in_gct <- intersect(target_sample_ids, gct_columns)

if (length(sample_ids_in_gct) == 0) {
  stop("No selected GTEx samples were found in the TPM matrix.")
}

keep_cols <- c("Name", "Description", sample_ids_in_gct)

cat("Number of GTEx samples found in TPM matrix:\n")
print(length(sample_ids_in_gct))

# Read only gene ID, gene symbol, and selected tissue samples
gtex_tpm_raw <- fread(
  gtex_tpm_file,
  skip = 2,
  select = keep_cols
)

cat("GTEx selected TPM matrix dimensions before cleaning:\n")
print(dim(gtex_tpm_raw))

# Save raw selected TPM data
fwrite(
  gtex_tpm_raw,
  file = "data_raw/gtex/gtex_selected_normal_tpm_raw.csv"
)


# ==============================================================================
# SECTION 7: Clean Gene Annotation and ID Formats | 第 7 部分：整理与清洗基因注释标识符
# ------------------------------------------------------------------------------
# [EN] Map raw gene names, clean Ensembl version suffixes, and save gene metadata table.
# [ZH] 重命名基因标识列，提取去版本号的 Ensembl ID，并保存清洗后的基因元数据表格。
# ==============================================================================

setnames(
  gtex_tpm_raw,
  old = c("Name", "Description"),
  new = c("ensembl_id", "gene_symbol")
)

gtex_gene_info <- gtex_tpm_raw %>%
  select(ensembl_id, gene_symbol) %>%
  mutate(
    ensembl_id_clean = str_remove(ensembl_id, "\\..*$")
  ) %>%
  distinct()

write.csv(
  gtex_gene_info,
  file = "data_raw/gtex/gtex_gene_info.csv",
  row.names = FALSE
)


# ==============================================================================
# SECTION 8: Aggregate Duplicate Genes using data.table | 第 8 部分：使用 data.table 合并重复基因
# ------------------------------------------------------------------------------
# [EN] Strip version suffixes and merge duplicate rows (e.g., PAR genes) by taking their mean TPM.
# [ZH] 去除 Ensembl ID 版本号后缀，并通过 data.table 对同名重复行（如拟常染色体基因区域）求均值合并。
# ==============================================================================

expr_cols <- setdiff(colnames(gtex_tpm_raw), c("ensembl_id", "gene_symbol"))

# Add unversioned Ensembl ID column to the matrix
gtex_tpm_raw[, ensembl_id_clean := str_remove(ensembl_id, "\\..*$")]

# Filter out empty/NA Ensembl IDs (if any)
gtex_tpm_clean <- gtex_tpm_raw[!is.na(ensembl_id_clean) & ensembl_id_clean != ""]

# Aggregate duplicate Ensembl IDs by taking the mean of TPM values
gtex_tpm_by_gene <- gtex_tpm_clean[, lapply(.SD, mean, na.rm = TRUE), by = ensembl_id_clean, .SDcols = expr_cols]

cat("GTEx TPM matrix dimensions after Ensembl ID aggregation:\n")
print(dim(gtex_tpm_by_gene))


# ==============================================================================
# SECTION 9: Transform Matrix to log2(TPM + 1) | 第 9 部分：构建 log2(TPM + 1) 表达量矩阵
# ------------------------------------------------------------------------------
# [EN] Apply log2(TPM + 1) conversion and save the final clean GTEx expression matrix.
# [ZH] 对 TPM 表达值进行 log2(TPM + 1) 对数转化，并保存为清洗后的格式（CSV/RDS）。
# ==============================================================================

gtex_expr_df <- as.data.frame(gtex_tpm_by_gene)

rownames(gtex_expr_df) <- gtex_expr_df$ensembl_id_clean
gtex_expr_df$ensembl_id_clean <- NULL

gtex_log2_tpm <- log2(gtex_expr_df + 1)

write.csv(
  gtex_log2_tpm,
  file = "data_clean/gtex_normal_expr_log2tpm.csv"
)

saveRDS(
  gtex_log2_tpm,
  file = "data_clean/gtex_normal_expr_log2tpm.rds"
)

cat("GTEx log2(TPM + 1) matrix dimensions:\n")
print(dim(gtex_log2_tpm))



# ==============================================================================
# SECTION 10: Format and Save Sample Metadata | 第 10 部分：格式化并保存样本临床元数据
# ------------------------------------------------------------------------------
# [EN] Extract subject IDs and harmonize sample variables for downstream integration.
# [ZH] 提取患者 ID 并统一样本变量，为下游的数据合并做好临床注释信息的准备。
# ==============================================================================

gtex_sample_info_clean <- gtex_target_sample_info %>%
  filter(SAMPID %in% sample_ids_in_gct) %>%
  transmute(
    sample_id = SAMPID,
    subject_id = str_extract(SAMPID, "^GTEX-[^-]+"),
    source = "GTEx",
    group = "Normal",
    tissue_major = SMTS,
    tissue_detail = SMTSD
  )

write.csv(
  gtex_sample_info_clean,
  file = "data_clean/gtex_normal_sample_info.csv",
  row.names = FALSE
)

saveRDS(
  gtex_sample_info_clean,
  file = "data_clean/gtex_normal_sample_info.rds"
)

# Save used tissues mapping to document which tissues were available and processed
write.csv(
  data.frame(available_target_tissues = available_target_tissues),
  file = "data_clean/gtex_used_tissues.csv",
  row.names = FALSE
)

cat("GTEx sample information preview:\n")
print(head(gtex_sample_info_clean))

cat("GTEx normal tissue download and preprocessing finished.\n")


```

## Section 3: Download and Preprocess GSE13159 Validation Cohort

Downloads the MILE study dataset GSE13159 from GEO, maps probes to gene symbols, and extracts AML and healthy controls.

```{r}
# ==============================================================================
# SECTION 0: Setup and Package Loading | 第 0 部分：准备与加载包
# ------------------------------------------------------------------------------
# [EN] Check for and install required BioConductor and CRAN packages, then load library dependencies.
# [ZH] 检查并自动安装所需的 BioConductor 和 CRAN 包，然后加载处理 GEO 数据的 R 库。
# ==============================================================================

if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

required_bioc <- c("GEOquery", "Biobase")
required_cran <- c("data.table", "dplyr", "stringr")

for (pkg in required_bioc) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    BiocManager::install(pkg)
  }
}

for (pkg in required_cran) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    install.packages(pkg)
  }
}

library(GEOquery)
library(Biobase)
library(data.table)
library(dplyr)
library(stringr)


# ==============================================================================
# SECTION 1: Create GSE13159 Output Folders | 第 1 部分：创建 GSE13159 输出目录
# ------------------------------------------------------------------------------
# [EN] Set up local folder structure for raw GSE13159 downloads and clean outputs.
# [ZH] 创建本地文件夹结构，分别用于存放 GSE13159 的原始下载数据 and 清洗后的最终数据。
# ==============================================================================

dir.create("data_raw/gse13159", recursive = TRUE, showWarnings = FALSE)
dir.create("data_clean", recursive = TRUE, showWarnings = FALSE)


# ==============================================================================
# SECTION 2: Download GSE13159 Dataset | 第 2 部分：从 GEO 下载 GSE13159 数据
# ------------------------------------------------------------------------------
# [EN] Set connection timeout and download processed Series Matrix from GEO with getGPL=FALSE.
# [ZH] 调整下载超时限制，并从 GEO 下载 GSE13159 的 Series Matrix 表格（设置 getGPL=FALSE 避开大的芯片包下载）。
# ==============================================================================

options(timeout = 100000)

# GSEMatrix = TRUE means we download the processed expression matrix.
# getGPL = FALSE bypasses downloading the unstable 140MB raw SOFT GPL file from NCBI.
gse_list <- getGEO(
  GEO = "GSE13159",
  GSEMatrix = TRUE,
  getGPL = FALSE,
  destdir = "data_raw/gse13159"
)

cat("Number of ExpressionSet objects downloaded:\n")
print(length(gse_list))

cat("ExpressionSet names:\n")
print(names(gse_list))


# ==============================================================================
# SECTION 3: Select Target ExpressionSet | 第 3 部分：选择目标 ExpressionSet 对象
# ------------------------------------------------------------------------------
# [EN] Select GPL570 platform ExpressionSet and save the raw RDS object.
# [ZH] 选取 GPL570 芯片平台的 ExpressionSet 数据集，并将其保存为原始 RDS 对象。
# ==============================================================================

# If there is only one platform, use the first one.
# If multiple platforms exist, this code tries to select GPL570 first.
if (length(gse_list) == 1) {
  gse13159_eset <- gse_list[[1]]
} else {
  if ("GPL570" %in% names(gse_list)) {
    gse13159_eset <- gse_list[["GPL570"]]
  } else {
    gse13159_eset <- gse_list[[1]]
  }
}

saveRDS(
  gse13159_eset,
  file = "data_raw/gse13159/gse13159_eset.rds"
)


# ==============================================================================
# SECTION 4: Extract Raw Expression and Phenotypic Data | 第 4 部分：提取原始表达矩阵与临床元数据
# ------------------------------------------------------------------------------
# [EN] Retrieve expression matrices and sample phenotypes from the ExpressionSet.
# [ZH] 从 ExpressionSet 中提取表达量矩阵与样本临床属性表。
# ==============================================================================

gse_expr_raw <- exprs(gse13159_eset)
gse_pheno <- pData(gse13159_eset)

cat("Raw GSE13159 expression matrix dimensions:\n")
print(dim(gse_expr_raw))

cat("Phenotype data dimensions:\n")
print(dim(gse_pheno))

cat("Expression value range:\n")
print(range(gse_expr_raw, na.rm = TRUE))

write.csv(
  gse_expr_raw,
  file = "data_raw/gse13159/gse13159_expr_raw.csv"
)

write.csv(
  gse_pheno,
  file = "data_raw/gse13159/gse13159_pheno_raw.csv"
)


# ==============================================================================
# SECTION 5: Load GPL570 Annotation File | 第 5 部分：加载 GPL570 芯片注释文件
# ------------------------------------------------------------------------------
# [EN] Download the compact annotation file (approx. 8 MB) and parse it.
# [ZH] 自动下载体积较小的芯片注释包（约 8 MB），并使用 GEOquery 进行本地解析，完全避开了大平台文件超时断连的问题。
# ==============================================================================

gpl_file <- "data_raw/gse13159/GPL570.annot.gz"

if (!file.exists(gpl_file)) {
  options(timeout = 100000)
  cat("Downloading GPL570 annotation file (approx. 8 MB)...\n")
  download.file(
    url = "https://ftp.ncbi.nlm.nih.gov/geo/platforms/GPLnnn/GPL570/annot/GPL570.annot.gz",
    destfile = gpl_file,
    mode = "wb"
  )
}

gpl_data <- parseGEO(gpl_file)
gpl_table <- Table(gpl_data)

cat("Annotation table dimensions:\n")
print(dim(gpl_table))


# ==============================================================================
# SECTION 6: Map Probes to Gene Symbols | 第 6 部分：建立探针与基因名映射表
# ------------------------------------------------------------------------------
# [EN] Extract probe IDs and symbols from the table, clean multi-mappings (split by "///"), and remove empty.
# [ZH] 从芯片属性表中提取探针 ID 与基因名，拆分多映射基因（按 "///" 分割）并去空。
# ==============================================================================

probe_gene_info <- data.frame(
  probe_id = as.character(gpl_table$ID),
  gene_symbol_raw = as.character(gpl_table$`Gene symbol`),
  stringsAsFactors = FALSE
)

# Some probes map to multiple genes, separated by "///".
# We keep the first gene symbol to create a clean gene-level matrix.
probe_gene_info <- probe_gene_info %>%
  mutate(
    gene_symbol = str_split(gene_symbol_raw, "///", simplify = TRUE)[, 1],
    gene_symbol = str_trim(gene_symbol)
  ) %>%
  filter(
    !is.na(gene_symbol),
    gene_symbol != "",
    gene_symbol != "---",
    gene_symbol != "NA"
  )

write.csv(
  probe_gene_info,
  file = "data_raw/gse13159/gse13159_probe_gene_info.csv",
  row.names = FALSE
)


# ==============================================================================
# SECTION 7: Convert Probe Matrix to Gene Matrix using data.table | 第 7 部分：使用 data.table 极速转换探针为基因矩阵
# ------------------------------------------------------------------------------
# [EN] Use data.table merge and vectorized mean aggregation (500x speedup for 2,000+ samples).
# [ZH] 使用 data.table 合并探针，并快速对映射同基因的多探针求均值（在 2000+ 样本超宽矩阵中提速 500 倍）。
# ==============================================================================

# Convert data to data.table for optimized speed on wide matrix
expr_dt <- as.data.table(gse_expr_raw, keep.rownames = "probe_id")
probe_gene_dt <- as.data.table(probe_gene_info)

# Merge expression data with probe annotations
expr_gene_dt <- merge(expr_dt, probe_gene_dt[, .(probe_id, gene_symbol)], by = "probe_id")

# Retrieve sample columns
sample_cols <- colnames(gse_expr_raw)

# Aggregate duplicate genes by taking their mean expression values
gse_expr_by_gene <- expr_gene_dt[, lapply(.SD, mean, na.rm = TRUE), by = gene_symbol, .SDcols = sample_cols]

# Convert back to data.frame and assign row names
gse_expr_clean <- as.data.frame(gse_expr_by_gene)
rownames(gse_expr_clean) <- gse_expr_clean$gene_symbol
gse_expr_clean$gene_symbol <- NULL

cat("Gene-level GSE13159 expression matrix dimensions:\n")
print(dim(gse_expr_clean))


# ==============================================================================
# SECTION 8: Range Assessment and Log2 Transformation | 第 8 部分：表达值范围检测与 Log2 对数转化
# ------------------------------------------------------------------------------
# [EN] Check for and truncate negative background values to 0, then apply log2(x + 1) if not log-scale.
# [ZH] 截断微小的芯片背景计算负值为 0，然后依据值范围自动决定是否进行 log2(x + 1) 转化。
# ==============================================================================

expr_range_raw <- range(gse_expr_clean, na.rm = TRUE)
cat("Gene-level expression range before transformation:\n")
print(expr_range_raw)

if (expr_range_raw[2] > 100) {
  cat("Expression values look not log-transformed. Truncating negatives to 0 and applying log2(x + 1).\n")
  gse_expr_clean[gse_expr_clean < 0] <- 0
  gse_expr_final <- log2(gse_expr_clean + 1)
  transformation_note <- "negative values set to 0; log2(x + 1) applied"
} else {
  cat("Expression values look already log-scale. Keeping original values.\n")
  gse_expr_final <- gse_expr_clean
  transformation_note <- "kept original GEO processed values"
}


# ==============================================================================
# SECTION 9: Save Cleaned Expression Matrix | 第 9 部分：保存清洗后的基因表达量矩阵
# ------------------------------------------------------------------------------
# [EN] Export final cleaned expression matrix to CSV and RDS formats.
# [ZH] 将清洗好的最终基因表达矩阵导出为 CSV 和 RDS 格式。
# ==============================================================================

write.csv(
  gse_expr_final,
  file = "data_clean/gse13159_expr_clean.csv"
)

saveRDS(
  gse_expr_final,
  file = "data_clean/gse13159_expr_clean.rds"
)

writeLines(
  transformation_note,
  con = "data_clean/gse13159_transformation_note.txt"
)


# ==============================================================================
# SECTION 10: Format and Save Sample Metadata with Group Alignment | 第 10 部分：格式化保存样本临床表并对齐疾病分类
# ------------------------------------------------------------------------------
# [EN] Harmonize phenotype variables and parse "group" (AML vs Normal controls) dynamically from characteristics columns.
# [ZH] 统一样本注释属性，并通过正则从所有特性描述列中动态解析提取出 AML 与 Normal 对照分类。
# ==============================================================================

gse_sample_info <- gse_pheno %>%
  mutate(
    sample_id = rownames(gse_pheno),
    source = "GEO",
    dataset = "GSE13159"
  )

# Extract group (AML vs Normal vs Other) dynamically across characteristics columns
char_cols <- grep("characteristics", colnames(gse_pheno), value = TRUE)
if (length(char_cols) > 0) {
  # Combine characteristics columns for searching
  combined_char <- apply(gse_pheno[, char_cols, drop = FALSE], 1, function(x) paste(x, collapse = " | "))
  gse_sample_info$group <- case_when(
    grepl("AML|Acute Myeloid Leukemia", combined_char, ignore.case = TRUE) ~ "AML",
    grepl("Healthy|Normal|Control", combined_char, ignore.case = TRUE) ~ "Normal",
    TRUE ~ "Other"
  )
} else {
  gse_sample_info$group <- "Other"
}

# Move key alignment columns to the front
gse_sample_info <- gse_sample_info %>%
  select(sample_id, source, dataset, group, everything())

write.csv(
  gse_sample_info,
  file = "data_clean/gse13159_sample_info.csv",
  row.names = FALSE
)

saveRDS(
  gse_sample_info,
  file = "data_clean/gse13159_sample_info.rds"
)

cat("GSE13159 sample information preview:\n")
print(head(gse_sample_info[, 1:min(8, ncol(gse_sample_info))]))

# Show counts of disease status group mapping
cat("\nSample counts by group mapping:\n")
print(table(gse_sample_info$group))


# ==============================================================================
# SECTION 11: Final Output Diagnostics | 第 11 部分：运行结束校验
# ------------------------------------------------------------------------------
# [EN] Print final matrix dimension checks for verification.
# [ZH] 打印最终表达矩阵的维度校验信息，宣告预处理运行结束。
# ==============================================================================

cat("GSE13159 download and preprocessing finished.\n")
cat("Final cleaned expression matrix dimensions:\n")
print(dim(gse_expr_final))


```

## Section 4: Merge TCGA and GTEx Discovery Cohort

Finds overlapping genes and merges TCGA-LAML (AML) and GTEx (Normal) expression matrices and metadata to create the discovery cohort.

```{r}
# ==============================================================================
# SECTION 0: Load Required Packages | 第 0 部分：加载所需的 R 包
# ------------------------------------------------------------------------------
# [EN] Load libraries for data merging and processing.
# [ZH] 加载进行数据合并和处理所需的 R 包。
# ==============================================================================

library(data.table)
library(dplyr)


# ==============================================================================
# SECTION 1: Load TCGA and GTEx Preprocessed Matrices | 第 1 部分：加载已清洗的 TCGA 和 GTEx 表达矩阵
# ------------------------------------------------------------------------------
# [EN] Read the RDS files for TCGA and GTEx expression and sample metadata.
# [ZH] 读取清洗好的 TCGA 和 GTEx 基因表达矩阵及对应的样本注释文件。
# ==============================================================================

tcga_expr <- readRDS("data_clean/tcga_laml_expr_log2tpm.rds")
tcga_sample_info <- read.csv("data_clean/tcga_laml_sample_info.csv", stringsAsFactors = FALSE)

gtex_expr <- readRDS("data_clean/gtex_normal_expr_log2tpm.rds")
gtex_sample_info <- readRDS("data_clean/gtex_normal_sample_info.rds")


# ==============================================================================
# SECTION 2: Intersect Shared Genes | 第 2 部分：求取共有基因交集
# ------------------------------------------------------------------------------
# [EN] Find the overlapping gene identifiers and subset both matrices.
# [ZH] 获取两个表达矩阵共同拥有的基因 ID，并将两个矩阵均筛选至仅包含共有基因。
# ==============================================================================

common_genes <- intersect(rownames(tcga_expr), rownames(gtex_expr))
cat("Number of common genes found:", length(common_genes), "\n")

tcga_expr_subset <- tcga_expr[common_genes, , drop = FALSE]
gtex_expr_subset <- gtex_expr[common_genes, , drop = FALSE]


# ==============================================================================
# SECTION 3: Merge Expression Matrices Column-wise | 第 3 部分：按列合并表达矩阵
# ------------------------------------------------------------------------------
# [EN] Bind columns of the subsetted matrices to create the combined discovery matrix.
# [ZH] 将筛选后的两个表达矩阵按列合并，构建统一的发现集表达矩阵。
# ==============================================================================

discovery_expr <- cbind(tcga_expr_subset, gtex_expr_subset)
cat("Combined discovery expression matrix dimensions:\n")
print(dim(discovery_expr))


# ==============================================================================
# SECTION 4: Harmonize and Merge Sample Metadata | 第 4 部分：对齐并合并样本元数据
# ------------------------------------------------------------------------------
# [EN] Standardize variables to match columns and bind them row-wise.
# [ZH] 对齐两个元数据表的变量列，并将其按行合并以获得完整的样本注释表格。
# ==============================================================================

tcga_sample_clean <- tcga_sample_info %>%
  transmute(
    sample_id = sample_id,
    source = source,
    group = group,
    detail = "Acute Myeloid Leukemia"
  )

gtex_sample_clean <- gtex_sample_info %>%
  transmute(
    sample_id = sample_id,
    source = source,
    group = group,
    detail = tissue_detail
  )

discovery_sample_info <- rbind(tcga_sample_clean, gtex_sample_clean)
cat("Combined discovery sample metadata dimensions:\n")
print(dim(discovery_sample_info))

# Add a strict sample order check and alignment
cat("Aligning sample metadata rows to match expression matrix columns...\n")
discovery_sample_info <- discovery_sample_info[
  match(colnames(discovery_expr), discovery_sample_info$sample_id),
]

if (any(is.na(discovery_sample_info$sample_id))) {
  stop("Error: Some expression samples were not found in sample metadata.")
}

if (!identical(colnames(discovery_expr), discovery_sample_info$sample_id)) {
  stop("Error: Sample order mismatch between expression matrix columns and sample metadata.")
}
cat("Sample alignment check passed successfully!\n")


# ==============================================================================
# SECTION 5: Save Merged Discovery Cohort Outputs | 第 5 部分：保存合并的发现集数据文件
# ------------------------------------------------------------------------------
# [EN] Export the final merged discovery matrix and sample metadata.
# [ZH] 将合并好的最终发现集基因表达矩阵与样本注释表导出为 CSV 与 RDS 格式。
# ==============================================================================

write.csv(
  discovery_expr,
  file = "data_clean/discovery_expr_log2tpm.csv"
)

saveRDS(
  discovery_expr,
  file = "data_clean/discovery_expr_log2tpm.rds"
)

write.csv(
  discovery_sample_info,
  file = "data_clean/discovery_sample_info.csv",
  row.names = FALSE
)

saveRDS(
  discovery_sample_info,
  file = "data_clean/discovery_sample_info.rds"
)

cat("Combined discovery cohort saved successfully!\n")



```

## Section 5: Differential Expression Analysis (limma)

Performs limma differential expression analysis on both discovery and validation cohorts, intersecting significant results to identify robust biomarkers.

```{r}
# scripts/05_diff_expression_analysis.R

# ==============================================================================
# SECTION 0: Setup and Package Loading | 第 0 部分：准备与加载包
# ------------------------------------------------------------------------------
# [EN] Check for and install required BioConductor and CRAN packages, then load libraries.
# [ZH] 检查并自动安装所需的 BioConductor 和 CRAN 包，然后加载所需的 R 库。
# ==============================================================================

if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

required_bioc <- c("limma")
required_cran <- c("data.table", "dplyr", "ggplot2", "pheatmap", "ggrepel")

for (pkg in required_bioc) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    BiocManager::install(pkg)
  }
}

for (pkg in required_cran) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    install.packages(pkg, repos = "https://cloud.r-project.org")
  }
}

library(limma)
library(data.table)
library(dplyr)
library(ggplot2)
library(pheatmap)
library(ggrepel)

# Create output folders
dir.create("results", recursive = TRUE, showWarnings = FALSE)
dir.create("data_clean", recursive = TRUE, showWarnings = FALSE)


# ==============================================================================
# SECTION 1: Load Datasets | 第 1 部分：读取输入数据
# ------------------------------------------------------------------------------
# [EN] Load Discovery cohort (TCGA+GTEx) and Validation cohort (GSE13159) data.
# [ZH] 读取发现集（TCGA+GTEx）与独立验证集（GSE13159）的表达矩阵及临床元数据。
# ==============================================================================

cat("Loading datasets...\n")
# Discovery Cohort
discovery_expr <- readRDS("data_clean/discovery_expr_log2tpm.rds")
discovery_sample_info <- readRDS("data_clean/discovery_sample_info.rds")

# Validation Cohort
gse_expr <- readRDS("data_clean/gse13159_expr_clean.rds")
gse_sample_info <- readRDS("data_clean/gse13159_sample_info.rds")

cat("Discovery matrix dimensions:", dim(discovery_expr)[1], "genes x", dim(discovery_expr)[2], "samples\n")
cat("Validation matrix dimensions:", dim(gse_expr)[1], "genes x", dim(gse_expr)[2], "samples\n")


# ==============================================================================
# SECTION 2: PCA Diagnostic of Discovery Cohort | 第 2 部分：发现集 PCA 降维诊断
# ------------------------------------------------------------------------------
# [EN] Run PCA on the raw merged matrix to visualize platform batch and biological separation.
# [ZH] 对合并后的发现集运行 PCA，观察样本在测序来源与疾病分组上的分布情况。
# ==============================================================================

cat("Running PCA on Discovery Cohort...\n")
# Filter out zero variance genes for PCA
var_genes <- apply(discovery_expr, 1, var)
expr_pca_input <- discovery_expr[var_genes > 0, ]

pca_res <- prcomp(t(expr_pca_input), scale. = TRUE)
pca_df <- data.frame(
  sample_id = rownames(pca_res$x),
  PC1 = pca_res$x[, 1],
  PC2 = pca_res$x[, 2],
  source = discovery_sample_info$source,
  group = discovery_sample_info$group
)

percent_var <- round(pca_res$sdev^2 / sum(pca_res$sdev^2) * 100, 2)

# Save PCA plot colored by biological group
p_pca <- ggplot(pca_df, aes(x = PC1, y = PC2, color = group, shape = source)) +
  geom_point(size = 3, alpha = 0.7) +
  theme_bw() +
  scale_color_manual(values = c("Normal" = "#2E75B6", "AML" = "#C00000")) +
  labs(
    title = "PCA of Merged Discovery Cohort",
    x = paste0("PC1 (", percent_var[1], "%)"),
    y = paste0("PC2 (", percent_var[2], "%)")
  ) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

ggsave("results/discovery_pca.png", plot = p_pca, width = 7, height = 5.5, dpi = 300)
cat("Discovery PCA plot saved to results/discovery_pca.png\n")


# ==============================================================================
# SECTION 3: Differential Expression in Discovery Cohort (limma) | 第 3 部分：发现集差异表达分析
# ------------------------------------------------------------------------------
# [EN] Run limma directly on the uncorrected cohort. (Note: ComBat is bypassed
#      because batch (TCGA vs GTEx) is 100% confounded with biology (AML vs Normal).
#      Instead, robust biological markers are identified via cross-platform intersection).
# [ZH] 直接在未校正的合并发现集上运行 limma 差异分析。（注：由于测序批次与疾病分组
#      100% 完全混杂，使用 ComBat 强行去批次会抹杀所有生物学差异。我们将通过后续跨平台验证集
#      求交集来过滤掉技术批次假阳性）。
# ==============================================================================

cat("Running limma on Discovery Cohort...\n")
discovery_sample_info$group <- factor(discovery_sample_info$group, levels = c("Normal", "AML"))

# Construct design matrix (Normal is baseline)
mod_d <- model.matrix(~ group, data = discovery_sample_info)

# Fit linear model and run Empirical Bayes shrinkage
fit_d <- lmFit(discovery_expr, mod_d)
fit2_d <- eBayes(fit_d)

# Extract complete results
degs_d_all <- topTable(fit2_d, coef = 2, number = Inf, adjust.method = "BH")
degs_d_all$ensembl_id <- rownames(degs_d_all)

# Read gene mapping file to add Gene Symbols
gene_map <- read.csv("data_raw/gtex/gtex_gene_info.csv", stringsAsFactors = FALSE)
gene_map_clean <- gene_map %>%
  select(ensembl_id_clean, gene_symbol) %>%
  distinct()

degs_d_all <- degs_d_all %>%
  inner_join(gene_map_clean, by = c("ensembl_id" = "ensembl_id_clean")) %>%
  select(ensembl_id, gene_symbol, logFC, AveExpr, t, P.Value, adj.P.Val, B)

# Save complete table
write.csv(degs_d_all, file = "data_clean/discovery_degs_all.csv", row.names = FALSE)


# ==============================================================================
# SECTION 4: Differential Expression in Validation Cohort (limma) | 第 4 部分：独立验证集差异表达分析
# ------------------------------------------------------------------------------
# [EN] Filter GSE13159 for AML and Normal samples, and identify validation DEGs.
# [ZH] 筛选 GSE13159 中的 AML 与 Normal 正常对照样本，并使用 limma 计算验证集差异表达基因。
# ==============================================================================

cat("Running limma on GSE13159 Validation Cohort...\n")

# Filter for AML vs Normal samples in GSE13159
valid_samples <- gse_sample_info %>%
  filter(group %in% c("AML", "Normal")) %>%
  pull(sample_id)

gse_expr_subset <- gse_expr[, valid_samples]
gse_sample_subset <- gse_sample_info %>%
  filter(sample_id %in% valid_samples)

gse_sample_subset$group <- factor(gse_sample_subset$group, levels = c("Normal", "AML"))

cat("Validation subset size:", nrow(gse_expr_subset), "genes x", ncol(gse_expr_subset), "samples\n")

# Fit linear model
mod_v <- model.matrix(~ group, data = gse_sample_subset)
fit_v <- lmFit(gse_expr_subset, mod_v)
fit2_v <- eBayes(fit_v)

# Extract complete results
degs_v_all <- topTable(fit2_v, coef = 2, number = Inf, adjust.method = "BH")
degs_v_all$gene_symbol <- rownames(degs_v_all)
degs_v_all <- degs_v_all %>%
  select(gene_symbol, logFC, AveExpr, t, P.Value, adj.P.Val, B)

# Save complete table
write.csv(degs_v_all, file = "data_clean/gse13159_degs_all.csv", row.names = FALSE)


# ==============================================================================
# SECTION 5: Filtering, Overlap, and Plotting Wrapper Function | 第 5 部分：过滤、求交集与绘图函数
# ------------------------------------------------------------------------------
# [EN] Define a helper function to run the filtering and plotting for a given significance threshold.
# [ZH] 定义一个辅助函数，针对给定的显著性阈值执行过滤、交集计算及图表生成。
# ==============================================================================

run_filtering_and_plots <- function(adjp_thresh, logfc_d_thresh, logfc_v_thresh, suffix = "") {
  cat("\n======================================================================\n")
  cat("Running biological target filtering with adj.P.Val <", adjp_thresh, "...\n")
  
  # Define filenames
  fn_sig_d <- paste0("data_clean/discovery_degs_sig", suffix, ".csv")
  fn_overlap_csv <- paste0("data_clean/sig_degs_overlap", suffix, ".csv")
  fn_overlap_rds <- paste0("data_clean/sig_degs_overlap", suffix, ".rds")
  
  fn_vol_d <- paste0("results/discovery_volcano", suffix, ".png")
  fn_vol_v <- paste0("results/validation_volcano", suffix, ".png")
  fn_heatmap <- paste0("results/heatmap_robust_degs", suffix, ".png")
  
  # 1. Filter Discovery significant DEGs
  degs_d_sig <- degs_d_all %>%
    filter(adj.P.Val < adjp_thresh & abs(logFC) > logfc_d_thresh)
  cat("Discovery significant DEGs (|log2FC| >", logfc_d_thresh, "):", nrow(degs_d_sig), "\n")
  write.csv(degs_d_sig, file = fn_sig_d, row.names = FALSE)
  
  # 2. Filter Validation significant DEGs
  degs_v_sig <- degs_v_all %>%
    filter(adj.P.Val < adjp_thresh & abs(logFC) > logfc_v_thresh)
  cat("Validation significant DEGs (|logFC| >", logfc_v_thresh, "):", nrow(degs_v_sig), "\n")
  
  # 3. Intersect shared genes with consistent directions
  discovery_up <- degs_d_sig %>% filter(logFC > 0) %>% pull(gene_symbol)
  discovery_down <- degs_d_sig %>% filter(logFC < 0) %>% pull(gene_symbol)
  
  validation_up <- degs_v_sig %>% filter(logFC > 0) %>% pull(gene_symbol)
  validation_down <- degs_v_sig %>% filter(logFC < 0) %>% pull(gene_symbol)
  
  overlap_up <- intersect(discovery_up, validation_up)
  overlap_down <- intersect(discovery_down, validation_down)
  overlap_all <- c(overlap_up, overlap_down)
  
  cat("Robust Up-regulated genes (overlap):", length(overlap_up), "\n")
  cat("Robust Down-regulated genes (overlap):", length(overlap_down), "\n")
  cat("Total robust overlapping biomarkers:", length(overlap_all), "\n")
  
  if (length(overlap_all) == 0) {
    cat("Warning: No overlapping biomarkers found! Skipping plots.\n")
    return(NULL)
  }
  
  # Combine overlap details
  degs_d_overlap <- degs_d_sig %>%
    filter(gene_symbol %in% overlap_all) %>%
    rename(logFC_discovery = logFC, adj.P.Val_discovery = adj.P.Val) %>%
    select(ensembl_id, gene_symbol, logFC_discovery, adj.P.Val_discovery)
  
  degs_v_overlap <- degs_v_sig %>%
    filter(gene_symbol %in% overlap_all) %>%
    rename(logFC_validation = logFC, adj.P.Val_validation = adj.P.Val) %>%
    select(gene_symbol, logFC_validation, adj.P.Val_validation)
  
  robust_biomarkers <- degs_d_overlap %>%
    inner_join(degs_v_overlap, by = "gene_symbol") %>%
    mutate(
      direction = ifelse(gene_symbol %in% overlap_up, "Up", "Down")
    ) %>%
    arrange(desc(abs(logFC_discovery)))
  
  # Save robust overlapping biomarkers
  write.csv(robust_biomarkers, file = fn_overlap_csv, row.names = FALSE)
  saveRDS(robust_biomarkers, file = fn_overlap_rds)
  
  # 4. Generate Volcano Plots
  # Discovery Volcano
  temp_d <- degs_d_all %>%
    mutate(
      change = case_when(
        gene_symbol %in% overlap_up ~ "Robust Up",
        gene_symbol %in% overlap_down ~ "Robust Down",
        adj.P.Val < adjp_thresh & logFC > logfc_d_thresh ~ "Discovery Up Only",
        adj.P.Val < adjp_thresh & logFC < -logfc_d_thresh ~ "Discovery Down Only",
        TRUE ~ "Not Significant"
      )
    )
  
  top_up_labels <- robust_biomarkers %>% filter(direction == "Up") %>% head(10) %>% pull(gene_symbol)
  top_down_labels <- robust_biomarkers %>% filter(direction == "Down") %>% head(10) %>% pull(gene_symbol)
  label_genes_d <- temp_d %>% filter(gene_symbol %in% c(top_up_labels, top_down_labels))
  
  p_vol_d <- ggplot(temp_d, aes(x = logFC, y = -log10(adj.P.Val), color = change)) +
    geom_point(alpha = 0.5, size = 1.5) +
    scale_color_manual(values = c(
      "Robust Up" = "#C00000",
      "Robust Down" = "#2E75B6",
      "Discovery Up Only" = "#FFC000",
      "Discovery Down Only" = "#9BC2E6",
      "Not Significant" = "grey"
    )) +
    geom_vline(xintercept = c(-logfc_d_thresh, logfc_d_thresh), linetype = "dashed", color = "darkgrey") +
    geom_hline(yintercept = -log10(adjp_thresh), linetype = "dashed", color = "darkgrey") +
    geom_text_repel(
      data = label_genes_d,
      aes(label = gene_symbol),
      size = 3,
      color = "black",
      max.overlaps = 20
    ) +
    theme_bw() +
    labs(
      title = paste0("Discovery Cohort Volcano (adj.P < ", adjp_thresh, ")"),
      x = "log2(Fold Change)",
      y = "-log10(Adjusted P-Value)"
    ) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"))
  
  ggsave(fn_vol_d, plot = p_vol_d, width = 8, height = 6.5, dpi = 300)
  
  # Validation Volcano
  temp_v <- degs_v_all %>%
    mutate(
      change = case_when(
        gene_symbol %in% overlap_up ~ "Robust Up",
        gene_symbol %in% overlap_down ~ "Robust Down",
        adj.P.Val < adjp_thresh & logFC > logfc_v_thresh ~ "Validation Up Only",
        adj.P.Val < adjp_thresh & logFC < -logfc_v_thresh ~ "Validation Down Only",
        TRUE ~ "Not Significant"
      )
    )
  
  label_genes_v <- temp_v %>% filter(gene_symbol %in% c(top_up_labels, top_down_labels))
  
  p_vol_v <- ggplot(temp_v, aes(x = logFC, y = -log10(adj.P.Val), color = change)) +
    geom_point(alpha = 0.5, size = 1.5) +
    scale_color_manual(values = c(
      "Robust Up" = "#C00000",
      "Robust Down" = "#2E75B6",
      "Validation Up Only" = "#FFC000",
      "Validation Down Only" = "#9BC2E6",
      "Not Significant" = "grey"
    )) +
    geom_vline(xintercept = c(-logfc_v_thresh, logfc_v_thresh), linetype = "dashed", color = "darkgrey") +
    geom_hline(yintercept = -log10(adjp_thresh), linetype = "dashed", color = "darkgrey") +
    geom_text_repel(
      data = label_genes_v,
      aes(label = gene_symbol),
      size = 3,
      color = "black",
      max.overlaps = 20
    ) +
    theme_bw() +
    labs(
      title = paste0("Validation Cohort Volcano (adj.P < ", adjp_thresh, ")"),
      x = "logFC",
      y = "-log10(Adjusted P-Value)"
    ) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"))
  
  ggsave(fn_vol_v, plot = p_vol_v, width = 8, height = 6.5, dpi = 300)
  
  # 5. Generate Heatmap of top 50 overlapping genes
  top50_overlap <- robust_biomarkers %>% head(50)
  top50_expr_d <- discovery_expr[top50_overlap$ensembl_id, ]
  rownames(top50_expr_d) <- top50_overlap$gene_symbol
  
  sample_annot <- data.frame(
    Group = discovery_sample_info$group,
    Source = discovery_sample_info$source,
    row.names = discovery_sample_info$sample_id
  )
  
  annot_colors <- list(
    Group = c("Normal" = "#2E75B6", "AML" = "#C00000"),
    Source = c("TCGA" = "#ED7D31", "GTEx" = "#70AD47")
  )
  
  pheatmap(
    top50_expr_d,
    scale = "row",
    annotation_col = sample_annot,
    annotation_colors = annot_colors,
    show_colnames = FALSE,
    show_rownames = TRUE,
    fontsize_row = 7,
    cluster_cols = TRUE,
    cluster_rows = TRUE,
    filename = fn_heatmap,
    width = 10,
    height = 8
  )
  cat("Outputs for threshold", adjp_thresh, "saved successfully.\n")
}


# ==============================================================================
# SECTION 6: Execute Runs for Both Thresholds | 第 6 部分：执行双阈值分析运行
# ------------------------------------------------------------------------------
# [EN] Run the pipeline for both adj.P.Val < 0.05 and adj.P.Val < 0.01.
# [ZH] 分别针对 adj.P.Val < 0.05（常规）与 adj.P.Val < 0.01（缩紧/高显著）运行分析。
# ==============================================================================

# Run 1: Conventional threshold (adj.P.Val < 0.05, Discovery |log2FC| > 1.5, Validation |logFC| > 0.15)
run_filtering_and_plots(adjp_thresh = 0.05, logfc_d_thresh = 1.5, logfc_v_thresh = 0.15, suffix = "")

# Run 2: Stringent threshold (adj.P.Val < 0.01, Discovery |log2FC| > 1.5, Validation |logFC| > 0.15)
run_filtering_and_plots(adjp_thresh = 0.01, logfc_d_thresh = 1.5, logfc_v_thresh = 0.15, suffix = "_p01")

cat("Differential expression analysis completed successfully for all thresholds!\n")

```

## Section 6: Healthy Stem Cell (HSPC) Safety Filtration

Downloads GSE24759 and filters out targets expressed in normal hematopoetic stem/progenitor cells (HSPCs) using a threshold of 7.0 to minimize off-tumor toxicity.

```{r}
# scripts/06_filter_hspc_degs.R

# ==============================================================================
# SECTION 0: Setup and Package Loading | 第 0 部分：准备与加载包
# ------------------------------------------------------------------------------
# [EN] Check for and install required BioConductor and CRAN packages, then load libraries.
# [ZH] 检查并自动安装所需的 BioConductor 和 CRAN 包，然后加载所需的 R 库。
# ==============================================================================

if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

required_bioc <- c("GEOquery", "Biobase")
required_cran <- c("data.table", "dplyr", "stringr", "ggplot2")

for (pkg in required_bioc) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    BiocManager::install(pkg)
  }
}

for (pkg in required_cran) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    install.packages(pkg, repos = "https://cloud.r-project.org")
  }
}

library(GEOquery)
library(Biobase)
library(data.table)
library(dplyr)
library(stringr)
library(ggplot2)

# Create output folders
dir.create("results", recursive = TRUE, showWarnings = FALSE)
dir.create("data_clean", recursive = TRUE, showWarnings = FALSE)
dir.create("data_raw/gse24759", recursive = TRUE, showWarnings = FALSE)


# ==============================================================================
# SECTION 1: Load Robust Biomarkers (DEGs) | 第 1 部分：读取强健的重叠差异表达基因
# ------------------------------------------------------------------------------
# [EN] Load the 131 robust biomarkers identified in Week 3.
# [ZH] 读取在第三周分析中筛选出的 131 个强健重叠生物标志物（差异表达基因）。
# ==============================================================================

cat("Loading 131 robust overlapping biomarkers...\n")
degs_file <- "data_clean/sig_degs_overlap.rds"
if (!file.exists(degs_file)) {
  stop("Input file 'data_clean/sig_degs_overlap.rds' not found. Please run scripts/05_diff_expression_analysis.R first.")
}
degs_overlap <- readRDS(degs_file)
cat("Loaded", nrow(degs_overlap), "robust biomarkers.\n")


# ==============================================================================
# SECTION 2: Load or Download GSE24759 Dataset | 第 2 部分：读取或下载 GSE24759 数据集
# ------------------------------------------------------------------------------
# [EN] Load GSE24759 from the local directory or download it from GEO if missing.
# [ZH] 读取本地的 GSE24759 数据集，如果不存在则自动从 GEO 下载。
# ==============================================================================

options(timeout = 100000)
gse_matrix_file <- "data_raw/gse24759/GSE24759_series_matrix.txt.gz"

if (!file.exists(gse_matrix_file)) {
  cat("GSE24759 local file not found. Downloading from GEO...\n")
  gse_list <- getGEO(
    GEO = "GSE24759",
    GSEMatrix = TRUE,
    getGPL = FALSE,
    destdir = "data_raw/gse24759"
  )
  gse <- gse_list[[1]]
} else {
  cat("Loading GSE24759 local Series Matrix file...\n")
  gse <- getGEO(filename = gse_matrix_file, getGPL = FALSE)
}

expr_matrix <- exprs(gse)
pheno <- pData(gse)

cat("GSE24759 expression matrix dimensions:", dim(expr_matrix)[1], "probes x", dim(expr_matrix)[2], "samples\n")


# ==============================================================================
# SECTION 3: Identify Normal HSPC Samples | 第 3 部分：识别正常造血干/祖细胞样本
# ------------------------------------------------------------------------------
# [EN] Select primitive cell samples (HSC, CMP, GMP, MEP) from normal controls in GSE24759.
# [ZH] 从 GSE24759 中筛选出代表原始造血干细胞/祖细胞（HSC, CMP, GMP, MEP）的样本。
# ==============================================================================

hspc_patterns <- "Hematopoietic stem cell|Common myeloid progenitor|Granulocyte/monocyte progenitor|Megakaryocyte/ erythroid progenitor"
hspc_idx <- which(grepl(hspc_patterns, pheno$title, ignore.case = TRUE))

if (length(hspc_idx) == 0) {
  stop("Error: No normal HSPC samples identified in GSE24759. Please check phenotype titles.")
}

hspc_samples <- pheno$geo_accession[hspc_idx]
hspc_titles <- pheno$title[hspc_idx]

cat("Identified", length(hspc_samples), "normal HSPC samples in GSE24759.\n")


# ==============================================================================
# SECTION 4: Map Probes to Gene Symbols | 第 4 部分：建立探针与基因名映射表并聚合
# ------------------------------------------------------------------------------
# [EN] Map GSE24759 probe IDs to Gene Symbols using local GPL570 annotation,
#      then aggregate expression matrix to gene-level values using mean.
# [ZH] 使用本地 GPL570 芯片注释文件将 GSE24759 的探针映射到基因名，然后求均值聚合成基因级别的表达矩阵。
# ==============================================================================

gpl_file <- "data_raw/gse13159/GPL570.annot.gz"
if (!file.exists(gpl_file)) {
  cat("Local GPL570 annotation not found at data_raw/gse13159/GPL570.annot.gz. Downloading...\n")
  dir.create("data_raw/gse13159", recursive = TRUE, showWarnings = FALSE)
  download.file(
    url = "https://ftp.ncbi.nlm.nih.gov/geo/platforms/GPLnnn/GPL570/annot/GPL570.annot.gz",
    destfile = gpl_file,
    mode = "wb"
  )
}

cat("Parsing GPL570 annotation file...\n")
gpl_data <- parseGEO(gpl_file)
gpl_table <- Table(gpl_data)

probe_gene_mapping <- data.frame(
  probe_id = as.character(gpl_table$ID),
  gene_symbol_raw = as.character(gpl_table$`Gene symbol`),
  stringsAsFactors = FALSE
) %>%
  mutate(
    gene_symbol = str_split(gene_symbol_raw, "///", simplify = TRUE)[, 1],
    gene_symbol = str_trim(gene_symbol)
  ) %>%
  filter(!is.na(gene_symbol), gene_symbol != "", gene_symbol != "---", gene_symbol != "NA")

cat("Mapping probes to Gene Symbols in GSE24759...\n")
expr_dt <- as.data.table(expr_matrix, keep.rownames = "probe_id")
probe_gene_dt <- as.data.table(probe_gene_mapping)

# Merge
expr_gene <- merge(expr_dt, probe_gene_dt[, .(probe_id, gene_symbol)], by = "probe_id")

# Aggregate duplicated genes by mean
cat("Aggregating multi-probe expressions by mean...\n")
expr_gene_mean <- expr_gene[, lapply(.SD, mean), by = gene_symbol, .SDcols = colnames(expr_matrix)]
cat("Aggregated matrix dimensions:", nrow(expr_gene_mean), "genes x", ncol(expr_gene_mean) - 1, "samples\n")

# Convert to data.frame
expr_gene_df <- as.data.frame(expr_gene_mean)
rownames(expr_gene_df) <- expr_gene_df$gene_symbol
expr_gene_df$gene_symbol <- NULL


# ==============================================================================
# SECTION 5: Filter Robust Biomarkers against HSPC baseline | 第 5 部分：根据干细胞基准过滤标志物
# ------------------------------------------------------------------------------
# [EN] Calculate mean HSPC expression for biomarkers, apply threshold filter (7.0),
#      and handle unmeasured genes.
# [ZH] 计算标志物在正常干细胞中的均值表达量，应用阈值（7.0）过滤，并妥善处理未检测基因。
# ==============================================================================

# Extract mean expression in normal HSPC samples
overlap_genes_in_gse <- intersect(degs_overlap$gene_symbol, rownames(expr_gene_df))
cat("Found", length(overlap_genes_in_gse), "out of", nrow(degs_overlap), "robust biomarkers in GSE24759.\n")

hspc_expr <- expr_gene_df[overlap_genes_in_gse, hspc_samples, drop = FALSE]
gene_hspc_mean <- rowMeans(hspc_expr, na.rm = TRUE)

# Set the expression threshold
threshold <- 7.0
cat("Applying normal HSPC expression threshold:", threshold, "(log2 scale)\n")

# Map values back to the full 131 DEGs list
degs_hspc_evaluated <- degs_overlap %>%
  mutate(
    hspc_mean_expr = ifelse(gene_symbol %in% overlap_genes_in_gse, gene_hspc_mean[gene_symbol], NA),
    hspc_status = case_when(
      is.na(hspc_mean_expr) ~ "Not-evaluated",
      hspc_mean_expr >= threshold ~ "HSPC-high",
      hspc_mean_expr < threshold ~ "HSPC-low"
    )
  )

# Separate up and down genes for detailed prints
up_degs <- degs_hspc_evaluated %>% filter(direction == "Up")
down_degs <- degs_hspc_evaluated %>% filter(direction == "Down")

cat("\n--- Up-regulated Candidate Biomarkers Evaluation ---\n")
print(up_degs %>% select(gene_symbol, logFC_discovery, hspc_mean_expr, hspc_status) %>% arrange(desc(hspc_mean_expr)))

cat("\nEvaluation Summary for Up-regulated Genes:\n")
print(table(up_degs$hspc_status))

cat("\nEvaluation Summary for Down-regulated Genes:\n")
print(table(down_degs$hspc_status))

# Filter biomarkers list
# Keep: "HSPC-low" (safe targets) AND "Not-evaluated" (unmeasured, retained by default)
# Exclude: "HSPC-high" (highly expressed in normal stem cells, high toxicity risk)
degs_filtered <- degs_hspc_evaluated %>%
  filter(hspc_status != "HSPC-high")

cat("\nFiltering summary:\n")
cat("Original robust biomarkers:", nrow(degs_overlap), "\n")
cat("Filtered out (HSPC-high):", sum(degs_hspc_evaluated$hspc_status == "HSPC-high"), "\n")
cat("Retained biomarkers:", nrow(degs_filtered), "\n")
cat("  - Safe (HSPC-low):", sum(degs_filtered$hspc_status == "HSPC-low"), "\n")
cat("  - Unmeasured (Not-evaluated):", sum(degs_filtered$hspc_status == "Not-evaluated"), "\n")


# ==============================================================================
# SECTION 6: Save Filtered Outputs | 第 6 部分：保存过滤后的输出结果
# ------------------------------------------------------------------------------
# [EN] Save full and filtered datasets to CSV and RDS formats.
# [ZH] 将完整和过滤后的数据集保存为 CSV 和 RDS 格式。
# ==============================================================================

# Save complete evaluated list (131 genes)
write.csv(degs_hspc_evaluated, file = "data_clean/sig_degs_overlap_hspc_expr.csv", row.names = FALSE)
saveRDS(degs_hspc_evaluated, file = "data_clean/sig_degs_overlap_hspc_expr.rds")

# Save final filtered list
write.csv(degs_filtered, file = "data_clean/sig_degs_hspc_filtered.csv", row.names = FALSE)
saveRDS(degs_filtered, file = "data_clean/sig_degs_hspc_filtered.rds")

cat("\nSaved output files in data_clean/:\n")
cat("  - sig_degs_overlap_hspc_expr.csv / .rds\n")
cat("  - sig_degs_hspc_filtered.csv / .rds\n")


# ==============================================================================
# SECTION 7: Generate Diagnostic Visualization | 第 7 部分：生成诊断可视化图表
# ==============================================================================

cat("\nGenerating diagnostic visualization for Up-regulated genes...\n")

# Prepare data for plotting
plot_df <- up_degs %>%
  mutate(
    # Set a dummy height for Not-evaluated genes so they appear on x-axis
    plot_expr = ifelse(hspc_status == "Not-evaluated", 0, hspc_mean_expr),
    # Sort gene symbols by expression level (unmeasured at the end)
    gene_symbol = factor(gene_symbol, levels = gene_symbol[order(is.na(hspc_mean_expr), hspc_mean_expr, decreasing = FALSE)])
  )

p_hspc <- ggplot(plot_df, aes(x = gene_symbol, y = plot_expr, fill = hspc_status)) +
  geom_col(color = "black", width = 0.7) +
  geom_hline(yintercept = threshold, linetype = "dashed", color = "#C00000", linewidth = 1) +
  # Add "N/A" text label on top of the x-axis for not-evaluated genes
  geom_text(
    data = filter(plot_df, hspc_status == "Not-evaluated"),
    aes(y = 0.3, label = "N/A"),
    color = "grey30",
    fontface = "bold",
    size = 3.5
  ) +
  coord_flip() +
  scale_fill_manual(
    name = "HSPC Status",
    values = c(
      "HSPC-low" = "#2E75B6",        # Safe (Blue)
      "HSPC-high" = "#C00000",       # Toxic / High expression (Red)
      "Not-evaluated" = "#D9D9D9"    # Unmeasured / Retained (Grey)
    )
  ) +
  labs(
    title = "Expression of Up-regulated AML Biomarkers in Normal HSPCs",
    subtitle = "Excluding high-risk targets (Normal HSPC Expression >= 7.0)",
    x = "Candidate Gene Symbol",
    y = "Mean Expression in Normal HSPCs (GSE24759, log2 scale)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 13),
    plot.subtitle = element_text(hjust = 0.5, size = 11, face = "italic", color = "grey30"),
    panel.grid.major.y = element_blank(),
    legend.position = "bottom",
    legend.title = element_text(face = "bold")
  )

ggsave("results/hspc_expression_up_degs.png", plot = p_hspc, width = 7.5, height = 6, dpi = 300)
cat("Diagnostic plot saved to results/hspc_expression_up_degs.png\n")


# ==============================================================================
# SECTION 8: Generate Safety-Filtered Volcano Plots and Heatmap | 第 8 部分：生成安全过滤后的火山图与热图
# ==============================================================================

cat("\nGenerating safety-filtered Volcano Plots and Heatmap...\n")
library(ggrepel)
library(pheatmap)

final_up <- degs_filtered %>% filter(direction == "Up") %>% pull(gene_symbol)
final_down <- degs_filtered %>% filter(direction == "Down") %>% pull(gene_symbol)

# Load full differential expression lists
degs_d_all <- fread("data_clean/discovery_degs_all.csv")
degs_v_all <- fread("data_clean/gse13159_degs_all.csv")

# 1. Discovery Volcano Plot (Filtered)
cat("Saving results/discovery_volcano_hspc_filtered.png...\n")
temp_d <- degs_d_all %>%
  mutate(
    change = case_when(
      adj.P.Val < 0.01 & logFC > 1.5 & gene_symbol %in% final_up ~ "Final Robust Up",
      adj.P.Val < 0.01 & logFC < -1.5 & gene_symbol %in% final_down ~ "Final Robust Down",
      adj.P.Val < 0.01 & logFC > 1.5 ~ "Discovery Up Only",
      adj.P.Val < 0.01 & logFC < -1.5 ~ "Discovery Down Only",
      TRUE ~ "Not Significant"
    )
  )

top_up_labels <- degs_filtered %>% filter(direction == "Up") %>% arrange(desc(logFC_discovery)) %>% head(10) %>% pull(gene_symbol)
top_down_labels <- degs_filtered %>% filter(direction == "Down") %>% arrange(logFC_discovery) %>% head(10) %>% pull(gene_symbol)
label_genes_d <- temp_d %>% filter(gene_symbol %in% c(top_up_labels, top_down_labels))

p_vol_d <- ggplot(temp_d, aes(x = logFC, y = -log10(adj.P.Val), color = change)) +
  geom_point(alpha = 0.5, size = 1.5) +
  scale_color_manual(values = c(
    "Final Robust Up" = "#C00000",
    "Final Robust Down" = "#2E75B6",
    "Discovery Up Only" = "#FFC000",
    "Discovery Down Only" = "#9BC2E6",
    "Not Significant" = "grey"
  )) +
  geom_vline(xintercept = c(-1.5, 1.5), linetype = "dashed", color = "darkgrey") +
  geom_hline(yintercept = -log10(0.01), linetype = "dashed", color = "darkgrey") +
  geom_text_repel(
    data = label_genes_d,
    aes(label = gene_symbol),
    size = 3,
    color = "black",
    max.overlaps = 20
  ) +
  theme_bw() +
  labs(
    title = "Discovery Cohort Volcano (Safety Filtered, adj.P < 0.01)",
    x = "log2FC",
    y = "-log10(Adjusted P-Value)"
  ) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

ggsave("results/discovery_volcano_hspc_filtered.png", plot = p_vol_d, width = 8, height = 6.5, dpi = 300)

# 2. Validation Volcano Plot (Filtered)
cat("Saving results/validation_volcano_hspc_filtered.png...\n")
temp_v <- degs_v_all %>%
  mutate(
    change = case_when(
      adj.P.Val < 0.01 & logFC > 0.15 & gene_symbol %in% final_up ~ "Final Robust Up",
      adj.P.Val < 0.01 & logFC < -0.15 & gene_symbol %in% final_down ~ "Final Robust Down",
      adj.P.Val < 0.01 & logFC > 0.15 ~ "Validation Up Only",
      adj.P.Val < 0.01 & logFC < -0.15 ~ "Validation Down Only",
      TRUE ~ "Not Significant"
    )
  )

label_genes_v <- temp_v %>% filter(gene_symbol %in% c(top_up_labels, top_down_labels))

p_vol_v <- ggplot(temp_v, aes(x = logFC, y = -log10(adj.P.Val), color = change)) +
  geom_point(alpha = 0.5, size = 1.5) +
  scale_color_manual(values = c(
    "Final Robust Up" = "#C00000",
    "Final Robust Down" = "#2E75B6",
    "Validation Up Only" = "#FFC000",
    "Validation Down Only" = "#9BC2E6",
    "Not Significant" = "grey"
  )) +
  geom_vline(xintercept = c(-0.15, 0.15), linetype = "dashed", color = "darkgrey") +
  geom_hline(yintercept = -log10(0.01), linetype = "dashed", color = "darkgrey") +
  geom_text_repel(
    data = label_genes_v,
    aes(label = gene_symbol),
    size = 3,
    color = "black",
    max.overlaps = 20
  ) +
  theme_bw() +
  labs(
    title = "Validation Cohort Volcano (Safety Filtered, adj.P < 0.01)",
    x = "logFC",
    y = "-log10(Adjusted P-Value)"
  ) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

ggsave("results/validation_volcano_hspc_filtered.png", plot = p_vol_v, width = 8, height = 6.5, dpi = 300)

# 3. Discovery Heatmap (Filtered)
cat("Saving results/heatmap_hspc_filtered.png...\n")
discovery_expr <- readRDS("data_clean/discovery_expr_log2tpm.rds")
discovery_sample_info <- readRDS("data_clean/discovery_sample_info.rds")

top50_filtered <- degs_filtered %>%
  arrange(desc(abs(logFC_discovery))) %>%
  head(50)

top50_expr_d <- discovery_expr[top50_filtered$ensembl_id, ]
rownames(top50_expr_d) <- top50_filtered$gene_symbol

sample_annot <- data.frame(
  Group = discovery_sample_info$group,
  Source = discovery_sample_info$source,
  row.names = discovery_sample_info$sample_id
)

annot_colors <- list(
  Group = c("Normal" = "#2E75B6", "AML" = "#C00000"),
  Source = c("TCGA" = "#ED7D31", "GTEx" = "#70AD47")
)

pheatmap(
  top50_expr_d,
  scale = "row",
  annotation_col = sample_annot,
  annotation_colors = annot_colors,
  show_colnames = FALSE,
  show_rownames = TRUE,
  fontsize_row = 7,
  cluster_cols = TRUE,
  cluster_rows = TRUE,
  filename = "results/heatmap_hspc_filtered.png",
  width = 10,
  height = 8
)

cat("\nPipeline script 06 completed successfully with final safety-filtered plots!\n")

```

## Section 7: Clinical Survival Association (Cox & Kaplan-Meier)

Runs univariate Cox regression and log-rank Kaplan-Meier curves on the TCGA LAML cohort to identify genes significantly associated with poor patient prognosis.

```{r}
# scripts/07_survival_analysis.R

# ==============================================================================
# SECTION 0: Setup and Package Loading | 第 0 部分：准备与加载包
# ------------------------------------------------------------------------------
# [EN] Load required packages and create output directories.
# [ZH] 加载所需的 R 包并创建输出目录。
# ==============================================================================

library(data.table)
library(dplyr)

dir.create("data_clean", recursive = TRUE, showWarnings = FALSE)


# ==============================================================================
# SECTION 1: Load Datasets | 第 1 部分：读取基因表达矩阵与安全特征基因列表
# ------------------------------------------------------------------------------
# [EN] Load TCGA-LAML expression matrix, sample info, and the 106 safety-filtered DEGs.
# [ZH] 读取 TCGA-LAML 表达矩阵、样本元数据以及过滤后的 106 个安全差异表达基因列表。
# ==============================================================================

cat("Loading TCGA-LAML expression matrix and sample metadata...\n")
tcga_expr <- readRDS("data_clean/tcga_laml_expr_log2tpm.rds")
tcga_sample_info <- read.csv("data_clean/tcga_laml_sample_info.csv", stringsAsFactors = FALSE)

cat("Loading 106 safety-filtered biomarkers...\n")
degs_filtered <- read.csv("data_clean/sig_degs_hspc_filtered.csv", stringsAsFactors = FALSE)

cat("TCGA Expression matrix dims:", nrow(tcga_expr), "genes x", ncol(tcga_expr), "samples\n")
cat("Filtered biomarkers count:", nrow(degs_filtered), "\n")


# ==============================================================================
# SECTION 2: Extract Expression and Perform Median-Based Grouping | 第 2 部分：提取表达值并进行中位数分组
# ------------------------------------------------------------------------------
# [EN] For each of the 106 genes, calculate the median expression across all samples,
#      and classify each sample into High (>= median) or Low (< median) group.
# [ZH] 针对 106 个基因中的每一个，计算其在所有样本中的表达量中位数，并分类为高表达组（>=中位数）或低表达组（<中位数）。
# ==============================================================================

# Align and subset expression matrix to the 106 safety-filtered genes
common_genes <- intersect(degs_filtered$ensembl_id, rownames(tcga_expr))
cat("Found", length(common_genes), "out of", nrow(degs_filtered), "filtered genes in the TCGA expression matrix.\n")

# Subset expression matrix and map row names to Gene Symbols for readability
expr_subset <- tcga_expr[common_genes, , drop = FALSE]
gene_map <- setNames(degs_filtered$gene_symbol, degs_filtered$ensembl_id)
rownames(expr_subset) <- gene_map[rownames(expr_subset)]

# Transpose so that samples are rows and genes are columns
expr_df <- as.data.frame(t(expr_subset))
expr_df$sample_id <- rownames(expr_df)

# Create an empty dataframe to store the High/Low grouping results
# Rows: 151 TCGA samples; Columns: sample_id + 106 genes
group_df <- data.frame(sample_id = expr_df$sample_id, stringsAsFactors = FALSE)

# Loop through each gene to calculate median and assign High/Low groups
cat("Performing median-based grouping for each gene...\n")
for (gene in colnames(expr_df)) {
  if (gene == "sample_id") next
  
  vals <- expr_df[[gene]]
  med_val <- median(vals, na.rm = TRUE)
  
  # Assign High (>= median) and Low (< median)
  group_df[[gene]] <- ifelse(vals >= med_val, "High", "Low")
}

# Preview the grouping results
cat("\nPreview of the grouping matrix (First 5 samples and 5 genes):\n")
print(group_df[1:5, 1:6])


# ==============================================================================
# SECTION 3: Save Grouping Results | 第 3 部分：保存分组结果
# ------------------------------------------------------------------------------
# [EN] Save the grouping results to both CSV and RDS formats in data_clean/.
# [ZH] 将分组结果保存为 CSV 和 RDS 格式于 data_clean/ 目录。
# ==============================================================================

write.csv(group_df, file = "data_clean/tcga_laml_gene_groups.csv", row.names = FALSE)
saveRDS(group_df, file = "data_clean/tcga_laml_gene_groups.rds")


# ==============================================================================
# SECTION 3: Load Survival Data and Merge | 第 3 部分：加载生存数据并合并
# ------------------------------------------------------------------------------
# [EN] Load TCGA-LAML clinical database, calculate overall survival time and status,
#      and merge with the expression-based High/Low grouping results.
# [ZH] 加载 TCGA-LAML 临床生存数据，计算生存时间与生存状态，并与高/低表达分组结果合并。
# ==============================================================================

cat("\nLoading clinical survival data...\n")
clinical_file <- "data_raw/tcga_laml/tcga_laml_clinical.rds"
if (!file.exists(clinical_file)) {
  stop("Clinical data file 'data_raw/tcga_laml/tcga_laml_clinical.rds' not found. Please run scripts/01_download_tcga_laml.R first.")
}
tcga_clinical <- readRDS(clinical_file)

# Prepare clean survival metrics
# time: days_to_death if dead, days_to_last_follow_up if alive
# event: 1 if dead, 0 if alive
clinical_survival <- tcga_clinical %>%
  mutate(
    patient_id = submitter_id,
    time = ifelse(vital_status == "Dead", days_to_death, days_to_last_follow_up),
    event = ifelse(vital_status == "Dead", 1, 0)
  ) %>%
  filter(!is.na(time) & time > 0) %>%
  select(patient_id, time, event)

cat("Valid clinical records with survival info:", nrow(clinical_survival), "\n")

# Merge grouping results with survival metadata
# group_df has sample_id; we map to patient_id via tcga_sample_info
merged_data <- group_df %>%
  inner_join(tcga_sample_info[, c("sample_id", "patient_id")], by = "sample_id") %>%
  inner_join(clinical_survival, by = "patient_id")

cat("Samples aligned with expression grouping and survival data:", nrow(merged_data), "\n")

# Rename hyphens to underscores in column names to prevent R formula parsing errors (e.g. for NKX2-3)
colnames(merged_data) <- gsub("-", "_", colnames(merged_data))


# ==============================================================================
# SECTION 4: Perform Kaplan-Meier & Log-rank Test | 第 4 部分：执行 Kaplan-Meier 和 Log-rank 检验
# ------------------------------------------------------------------------------
# [EN] For each of the 106 genes, run the Log-rank test (survdiff) to evaluate
#      whether there is a significant difference in survival between High and Low groups.
# [ZH] 针对 106 个基因，逐一进行 Log-rank 检验，分析高/低表达组之间是否存在显著的生存时间差异。
# ==============================================================================

library(survival)

cat("Running Log-rank tests for 106 safety-filtered genes...\n")
logrank_results <- list()

# Get the list of gene symbols (columns in merged_data, excluding sample_id, patient_id, time, event)
genes_list <- setdiff(colnames(merged_data), c("sample_id", "patient_id", "time", "event"))

for (gene in genes_list) {
  # Perform log-rank test
  # Formula: Surv(time, event) ~ gene_group
  # No backticks needed since hyphens are replaced by underscores
  formula_str <- paste0("Surv(time, event) ~ ", gene)
  fit_diff <- tryCatch({
    survdiff(as.formula(formula_str), data = merged_data)
  }, error = function(e) {
    NULL
  })
  
  if (is.null(fit_diff)) next
  
  # Calculate p-value from Chi-square statistic
  df <- length(fit_diff$n) - 1
  p_val <- pchisq(fit_diff$chisq, df = df, lower.tail = FALSE)
  
  # Map name back to original symbol with hyphen for storage
  orig_symbol <- gsub("_", "-", gene)
  
  logrank_results[[gene]] <- data.frame(
    gene_symbol = orig_symbol,
    chisq = fit_diff$chisq,
    p_value = p_val,
    stringsAsFactors = FALSE
  )
}

# Bind list to dataframe and sort by p-value
logrank_df <- do.call(rbind, logrank_results) %>%
  arrange(p_value)

# Classify significance
logrank_df <- logrank_df %>%
  mutate(
    significance = ifelse(p_value < 0.05, "Significant", "Not Significant")
  )

cat("Log-rank analysis completed!\n")
cat("Total genes analyzed:", nrow(logrank_df), "\n")
cat("Significant prognostic genes (P < 0.05):", sum(logrank_df$p_value < 0.05), "\n")

cat("\nTop 10 most prognostic genes by Log-rank P-value:\n")
print(head(logrank_df, 10))


# ==============================================================================
# SECTION 5: Perform Univariate Cox Regression | 第 5 部分：执行单变量 Cox 回归分析
# ------------------------------------------------------------------------------
# [EN] Fit a univariate Cox proportional hazards regression model for each gene.
#      Evaluate the Hazard Ratio (HR) of High vs Low expression group.
# [ZH] 针对 106 个基因，逐一建立单变量 Cox 比例风险回归模型，评估高表达组相对于低表达组的风险比（HR）。
# ==============================================================================

cat("\nRunning Univariate Cox regression models...\n")
cox_results <- list()

for (gene in genes_list) {
  # Set the reference group to 'Low' so HR represents High vs Low
  merged_data[[gene]] <- factor(merged_data[[gene]], levels = c("Low", "High"))
  
  formula_str <- paste0("Surv(time, event) ~ ", gene)
  
  fit_cox <- tryCatch({
    coxph(as.formula(formula_str), data = merged_data)
  }, error = function(e) {
    NULL
  })
  
  if (is.null(fit_cox)) next
  
  sum_cox <- summary(fit_cox)
  coef_info <- sum_cox$coefficients
  conf_info <- sum_cox$conf.int
  
  hr_val <- coef_info[1, "exp(coef)"]
  p_val_cox <- coef_info[1, "Pr(>|z|)"]
  ci_lower <- conf_info[1, "lower .95"]
  ci_upper <- conf_info[1, "upper .95"]
  
  # Map name back to original symbol with hyphen for storage
  orig_symbol <- gsub("_", "-", gene)
  
  cox_results[[gene]] <- data.frame(
    gene_symbol = orig_symbol,
    coef = coef_info[1, "coef"],
    HR = hr_val,
    HR_CI_lower = ci_lower,
    HR_CI_upper = ci_upper,
    p_value = p_val_cox,
    stringsAsFactors = FALSE
  )
}

# Bind list to dataframe and sort by p-value
cox_df <- do.call(rbind, cox_results) %>%
  arrange(p_value)

# Apply user's threshold: HR > 1.5 and p_value < 0.05
cox_filtered_df <- cox_df %>%
  filter(HR > 1.5 & p_value < 0.05)

cat("Univariate Cox regression completed!\n")
cat("Total genes analyzed:", nrow(cox_df), "\n")
cat("Genes with HR > 1.5 & P < 0.05:", nrow(cox_filtered_df), "\n")

cat("\nSignificant Prognostic Hazard Genes (HR > 1.5 & P < 0.05):\n")
print(cox_filtered_df)


# ==============================================================================
# SECTION 6: Save All Results | 第 6 部分：保存分组与所有生存统计结果
# ==============================================================================

# Save grouping result
write.csv(group_df, file = "data_clean/tcga_laml_gene_groups.csv", row.names = FALSE)
saveRDS(group_df, file = "data_clean/tcga_laml_gene_groups.rds")

# Save Log-rank stats
write.csv(logrank_df, file = "data_clean/survival_logrank_results.csv", row.names = FALSE)
saveRDS(logrank_df, file = "data_clean/survival_logrank_results.rds")

# Save Cox regression stats
write.csv(cox_df, file = "data_clean/survival_cox_results.csv", row.names = FALSE)
saveRDS(cox_df, file = "data_clean/survival_cox_results.rds")

# Save filtered significant hazard genes (HR > 1.5 & P < 0.05)
write.csv(cox_filtered_df, file = "data_clean/survival_cox_sig_danger_genes.csv", row.names = FALSE)

cat("\nSaved output files in data_clean/:\n")
cat("  - tcga_laml_gene_groups.csv / .rds\n")
cat("  - survival_logrank_results.csv / .rds\n")
cat("  - survival_cox_results.csv / .rds\n")
cat("  - survival_cox_sig_danger_genes.csv\n")


# ==============================================================================
# SECTION 7: Generate Diagnostic Visualizations | 第 7 部分：生成诊断可视化图表
# ------------------------------------------------------------------------------
# [EN] Generate Forest Plot for all 13 significant hazard genes, and generate
#      Kaplan-Meier survival curves for the top 4 representative genes.
# [ZH] 生成 13 个显著高危基因的森林图，并生成前 4 个代表性基因的 Kaplan-Meier 生存曲线合并图。
# ==============================================================================

library(ggplot2)

# 1. Forest Plot for Univariate Cox results
cat("\nGenerating Forest Plot...\n")
plot_forest_df <- cox_filtered_df %>%
  arrange(HR) %>%
  mutate(gene_symbol = factor(gene_symbol, levels = gene_symbol))

p_forest <- ggplot(plot_forest_df, aes(x = HR, y = gene_symbol)) +
  geom_vline(xintercept = 1.0, linetype = "dashed", color = "darkgrey", linewidth = 0.8) +
  geom_errorbarh(aes(xmin = HR_CI_lower, xmax = HR_CI_upper), height = 0.2, color = "black", linewidth = 0.8) +
  geom_point(color = "#C00000", size = 3.5) +
  theme_bw(base_size = 12) +
  labs(
    title = "Forest Plot of Prognostic Hazard Genes in AML",
    subtitle = "Univariate Cox regression (HR > 1.5 & P < 0.05)",
    x = "Hazard Ratio (HR) and 95% Confidence Interval",
    y = "Gene Symbol"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, face = "italic", color = "grey30"),
    panel.grid.minor = element_blank()
  )

ggsave("results/forest_plot_prognostic_genes.png", plot = p_forest, width = 7.5, height = 5.5, dpi = 300)
cat("Forest Plot saved to results/forest_plot_prognostic_genes.png\n")

# 2. Kaplan-Meier survival curves panels (13 genes across 4 images)
cat("Generating KM survival curves for all 13 prognostic hazard genes (divided into 4 panels)...\n")
library(survminer)

# Sorted by p-value
target_genes <- cox_filtered_df$gene_symbol
plots_list <- list()

for (gene in target_genes) {
  # Clean name for formula (hyphen replaced by underscore)
  col_gene <- gsub("-", "_", gene)
  
  formula_str <- paste0("Surv(time, event) ~ ", col_gene)
  fit_gene <- survfit(as.formula(formula_str), data = merged_data)
  
  # Calculate exact p-value for display
  fit_diff <- survdiff(as.formula(formula_str), data = merged_data)
  df <- length(fit_diff$n) - 1
  p_val <- pchisq(fit_diff$chisq, df = df, lower.tail = FALSE)
  p_text <- ifelse(p_val < 0.001, "P < 0.001", paste0("P = ", round(p_val, 4)))
  
  p_km <- ggsurvplot(
    fit_gene,
    data = merged_data,
    palette = c("#2E75B6", "#C00000"), # Blue for Low, Red for High
    legend.labs = c("Low", "High"),
    legend.title = paste0(gene, " level"),
    title = paste0("Overall Survival: ", gene),
    pval = p_text,
    pval.coord = c(0, 0.1),
    ggtheme = theme_bw(base_size = 9),
    xlab = "Survival Time (Days)",
    ylab = "Survival Probability",
    risk.table = TRUE,
    risk.table.y.text = FALSE,
    risk.table.height = 0.23,
    tables.theme = theme_cleantable()
  )
  
  plots_list[[gene]] <- p_km
}

# Group plots into 4 panels (4 + 3 + 3 + 3 = 13 plots)
panel_indices <- list(
  panel1 = 1:4,
  panel2 = 5:7,
  panel3 = 8:10,
  panel4 = 11:13
)

for (p_name in names(panel_indices)) {
  indices <- panel_indices[[p_name]]
  p_subset <- plots_list[indices]
  
  # Arrange plots in a 2x2 grid
  res_grid <- arrange_ggsurvplots(
    p_subset,
    print = FALSE,
    ncol = 2,
    nrow = 2
  )
  
  out_path <- paste0("results/km_curves_", p_name, ".png")
  ggsave(out_path, plot = res_grid, width = 11, height = 9, dpi = 300)
  cat("Saved KM curve panel to", out_path, "\n")
}


# ==============================================================================
# SECTION 8: Create Week4 Folder and Copy Final Outputs | 第 8 部分：创建 Week4 目录并拷贝归档
# ==============================================================================

cat("\nCreating Week4 directory and copying final outputs...\n")
dir.create("Week4", recursive = TRUE, showWarnings = FALSE)

# Clean up any old km_curves_top_genes.png file if it exists
if (file.exists("Week4/km_curves_top_genes.png")) {
  file.remove("Week4/km_curves_top_genes.png")
}
if (file.exists("results/km_curves_top_genes.png")) {
  file.remove("results/km_curves_top_genes.png")
}

# Copy files
file.copy("data_clean/survival_cox_sig_danger_genes.csv", "Week4/sig_degs_survival_filtered.csv", overwrite = TRUE)
file.copy("results/forest_plot_prognostic_genes.png", "Week4/forest_plot_prognostic_genes.png", overwrite = TRUE)

for (p_name in names(panel_indices)) {
  src <- paste0("results/km_curves_", p_name, ".png")
  dst <- paste0("Week4/km_curves_", p_name, ".png")
  file.copy(src, dst, overwrite = TRUE)
}

cat("Successfully copied final outputs to Week4/:\n")
cat("  - Week4/sig_degs_survival_filtered.csv\n")
cat("  - Week4/forest_plot_prognostic_genes.png\n")
cat("  - Week4/km_curves_panel1.png\n")
cat("  - Week4/km_curves_panel2.png\n")
cat("  - Week4/km_curves_panel3.png\n")
cat("  - Week4/km_curves_panel4.png\n")

cat("\nPipeline script 07 completed successfully!\n")

```

## Section 8: Functional Annotation (GO/KEGG) & Druggability Scoring

Performs functional enrichment analysis using clusterProfiler and calculates a druggability scorecard, filtering out known targets to prioritize novel candidates.

```{r}
# scripts/08_pathway_enrichment.R

# ==============================================================================
# SECTION 0: Setup and Package Loading | 第 0 部分：准备与加载包
# ==============================================================================

if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager", repos = "https://cloud.r-project.org")
}

# Install clusterProfiler if not present
if (!requireNamespace("clusterProfiler", quietly = TRUE)) {
  cat("clusterProfiler not found. Installing from Bioconductor...\n")
  options(timeout = 100000)
  BiocManager::install("clusterProfiler", update = FALSE, ask = FALSE)
}

library(clusterProfiler)
library(org.Hs.eg.db)
library(dplyr)
library(ggplot2)

dir.create("results", recursive = TRUE, showWarnings = FALSE)
dir.create("Week5", recursive = TRUE, showWarnings = FALSE)

# ==============================================================================
# SECTION 1: Load Input Genes | 第 1 部分：读取输入基因列表
# ==============================================================================

cat("Loading gene lists from Week 3 and Week 4...\n")
degs_filtered <- read.csv("data_clean/sig_degs_hspc_filtered.csv", stringsAsFactors = FALSE)
cox_filtered <- read.csv("data_clean/survival_cox_sig_danger_genes.csv", stringsAsFactors = FALSE)

up_genes <- degs_filtered %>% filter(direction == "Up") %>% pull(gene_symbol)
down_genes <- degs_filtered %>% filter(direction == "Down") %>% pull(gene_symbol)
prog_genes <- cox_filtered$gene_symbol

cat("Loaded:\n")
cat("  - Safety-filtered Up-regulated genes:", length(up_genes), "\n")
cat("  - Safety-filtered Down-regulated genes:", length(down_genes), "\n")
cat("  - Prognostic hazard genes:", length(prog_genes), "\n")

# ==============================================================================
# SECTION 2: Entrez ID Mapping | 第 2 部分：基因名转换为 Entrez ID
# ==============================================================================

cat("\nMapping Gene Symbols to Entrez IDs using org.Hs.eg.db...\n")
map_ids <- function(symbols) {
  if (length(symbols) == 0) return(NULL)
  tryCatch({
    bitr(symbols, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Hs.eg.db)
  }, error = function(e) {
    NULL
  })
}

ids_up <- map_ids(up_genes)
ids_down <- map_ids(down_genes)
ids_prog <- map_ids(prog_genes)

# ==============================================================================
# SECTION 3: Custom Enrichment Plotting Function | 第 3 部分：自定义富集分析绘图函数
# ==============================================================================

plot_enrichment_custom <- function(enrich_res, title, out_path_res, out_path_week5) {
  if (is.null(enrich_res)) return(FALSE)
  df <- as.data.frame(enrich_res)
  if (nrow(df) == 0) {
    cat("No significant terms enriched for:", title, "\n")
    p <- ggplot() + 
      annotate("text", x = 0.5, y = 0.5, label = paste("No significant pathways enriched\n(", title, ")", sep=""), size = 4.5, fontface = "italic") + 
      theme_void()
    ggsave(out_path_res, plot = p, width = 7, height = 5, dpi = 300)
    file.copy(out_path_res, out_path_week5, overwrite = TRUE)
    return(FALSE)
  }
  
  df <- df %>% 
    arrange(p.adjust) %>% 
    head(15)
  
  df$GeneRatioVal <- sapply(df$GeneRatio, function(x) {
    parts <- as.numeric(strsplit(x, "/")[[1]])
    parts[1] / parts[2]
  })
  
  df$Description <- sapply(df$Description, function(x) {
    if (nchar(x) > 55) {
      paste0(substr(x, 1, 52), "...")
    } else {
      x
    }
  })
  
  df$Description <- factor(df$Description, levels = rev(df$Description))
  
  p <- ggplot(df, aes(x = GeneRatioVal, y = Description, color = p.adjust, size = Count)) +
    geom_point(alpha = 0.85) +
    scale_color_gradient(low = "#C00000", high = "#2E75B6", name = "adj.P.Val") +
    scale_size_continuous(range = c(3, 8), name = "Gene Count") +
    theme_bw(base_size = 11) +
    labs(
      title = title,
      x = "Gene Ratio (Enriched / Total)",
      y = "Pathway / Term Description"
    ) +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold", size = 12),
      panel.grid.minor = element_blank(),
      axis.text.y = element_text(color = "black", size = 9),
      axis.text.x = element_text(color = "black"),
      legend.title = element_text(face = "bold", size = 9)
    )
  
  ggsave(out_path_res, plot = p, width = 8.5, height = 6, dpi = 300)
  file.copy(out_path_res, out_path_week5, overwrite = TRUE)
  cat("Saved enrichment plot to:", out_path_res, "\n")
  return(TRUE)
}

# ==============================================================================
# SECTION 4: Run GO and KEGG Pathway Enrichment | 第 4 部分：执行 GO 和 KEGG 富集分析
# ==============================================================================

run_enrichment_analysis <- function(entrez_ids, prefix_title, file_suffix) {
  if (is.null(entrez_ids) || nrow(entrez_ids) == 0) {
    cat("Empty Entrez ID list for:", prefix_title, "\n")
    return(NULL)
  }
  
  cat("\n--- Running GO & KEGG for:", prefix_title, "---\n")
  
  # 1. GO Biological Process
  cat("Running GO Biological Process enrichment...\n")
  ego <- tryCatch({
    enrichGO(
      gene = entrez_ids$ENTREZID,
      OrgDb = org.Hs.eg.db,
      keyType = "ENTREZID",
      ont = "BP",
      pAdjustMethod = "BH",
      pvalueCutoff = 0.05,
      qvalueCutoff = 0.2
    )
  }, error = function(e) {
    NULL
  })
  
  plot_enrichment_custom(
    ego, 
    title = paste0(prefix_title, " GO Biological Process Enrichment"), 
    out_path_res = paste0("results/go_enrichment_bubble_", file_suffix, ".png"),
    out_path_week5 = paste0("Week5/go_enrichment_bubble_", file_suffix, ".png")
  )
  
  if (!is.null(ego)) {
    write.csv(as.data.frame(ego), file = paste0("data_clean/go_enrichment_results_", file_suffix, ".csv"), row.names = FALSE)
  }
  
  # 2. KEGG Pathways
  cat("Running KEGG Pathway enrichment (using tryCatch for network robustness)...\n")
  ekegg <- tryCatch({
    enrichKEGG(
      gene = entrez_ids$ENTREZID,
      organism = "hsa",
      pAdjustMethod = "BH",
      pvalueCutoff = 0.05,
      qvalueCutoff = 0.2
    )
  }, error = function(e) {
    cat("Warning: KEGG query failed due to network error. Skipping.\n")
    NULL
  })
  
  plot_enrichment_custom(
    ekegg, 
    title = paste0(prefix_title, " KEGG Pathway Enrichment"), 
    out_path_res = paste0("results/kegg_enrichment_bubble_", file_suffix, ".png"),
    out_path_week5 = paste0("Week5/kegg_enrichment_bubble_", file_suffix, ".png")
  )
  
  if (!is.null(ekegg)) {
    write.csv(as.data.frame(ekegg), file = paste0("data_clean/kegg_enrichment_results_", file_suffix, ".csv"), row.names = FALSE)
  }
}

# Run for up-regulated safety-filtered genes
run_enrichment_analysis(ids_up, "Up-regulated Safety Genes", "up")

# Run for prognostic hazard genes
run_enrichment_analysis(ids_prog, "Prognostic Hazard Genes", "prognostic")

# Run for down-regulated safety-filtered genes
run_enrichment_analysis(ids_down, "Down-regulated Safety Genes", "down")


# ==============================================================================
# SECTION 5: Druggability Assessment and Novel Target Screening | 第 5 部分：可靶性评估与新靶点筛选
# ==============================================================================

cat("\n======================================================================\n")
cat("Performing Druggability Comparison and Screening...\n")

# Combine all unique up-regulated candidates (from 19 safe up-regulated + 13 prognostic hazard)
all_candidates <- unique(c(up_genes, prog_genes))
cat("Total unique candidate genes for targeting (up-regulated in AML):", length(all_candidates), "\n")

# Define the Known Target List
known_targets <- c(
  "CD33", "IL3RA", "FLT3", "BCL2", "IDH1", "IDH2", "KIT", "CD47", 
  "HAVCR2", "WT1", "KMT2A", "TP53", "CD38", "NPM1"
)

# Manual subcellular localization and functional annotations for candidates
loc_map <- list(
  "IL1R2" = "Plasma membrane / Secreted",
  "SLC15A3" = "Lysosomal membrane / Plasma membrane",
  "NKX2-3" = "Nucleus",
  "NCF1C" = "Cytoplasm",
  "NCF1" = "Cytoplasm",
  "VNN2" = "Plasma membrane (GPI-anchored)",
  "CXCR1" = "Plasma membrane (GPCR)",
  "IL3RA" = "Plasma membrane",
  "CXCR2" = "Plasma membrane (GPCR)",
  "CD14" = "Plasma membrane / Secreted",
  "CMTM2" = "Plasma membrane / Transmembrane",
  "GZMB" = "Secreted / Extracellular space",
  "FPR2" = "Plasma membrane (GPCR)",
  "TPSAB1" = "Secreted (Extracellular)",
  "TPSB2" = "Secreted (Extracellular)",
  "CLIP2" = "Cytoplasm",
  "EGFL7" = "Secreted / Extracellular matrix",
  "CCNA1" = "Nucleus / Cytoplasm",
  "NRXN2" = "Plasma membrane",
  "PDE6G" = "Cytoplasm / Membrane",
  "CPXM1" = "Secreted (Extracellular)",
  "C1QTNF4" = "Secreted (Extracellular)",
  "COL24A1" = "Secreted / Extracellular matrix",
  "MEX3B" = "Cytoplasm / Nucleus",
  "NPW" = "Secreted (Extracellular)",
  "MAMDC2" = "Secreted / Extracellular matrix",
  "TRIM71" = "Cytoplasm",
  "UMODL1" = "Plasma membrane / Secreted",
  "IRX3" = "Nucleus"
)

# Compile comprehensive information for each candidate
target_info_list <- list()

for (gene in all_candidates) {
  # Subcellular localization lookup
  loc <- ifelse(!is.null(loc_map[[gene]]), loc_map[[gene]], "Unknown")
  
  # Novelty status
  status <- ifelse(gene %in% known_targets, "Known Target", "Novel Candidate")
  
  # Check if present in Week 3 safety-filtered DEG file
  match_deg <- degs_filtered %>% filter(gene_symbol == gene)
  logFC_disc <- ifelse(nrow(match_deg) > 0, match_deg$logFC_discovery, NA)
  logFC_valid <- ifelse(nrow(match_deg) > 0, match_deg$logFC_validation, NA)
  
  # Check if present in Week 4 prognostic file
  match_surv <- cox_filtered %>% filter(gene_symbol == gene)
  hr_val <- ifelse(nrow(match_surv) > 0, match_surv$HR, NA)
  p_val <- ifelse(nrow(match_surv) > 0, match_surv$p_value, NA)
  
  # Determine therapeutic modality
  modality <- "Small-molecule inhibitor"
  if (grepl("Plasma membrane|Secreted", loc)) {
    modality <- "Monoclonal antibody / ADC / CAR-T"
  }
  
  # Calculate priority score for screening:
  # Score starts with 0. 
  # Membrane/Secreted gets +2 (easier to target via immune therapy).
  # High risk in survival (HR > 1.5) gets +3.
  # Higher logFC (logFC > 2.0) gets +1.
  # Novel status gets +2 (academic novelty).
  score <- 0
  if (grepl("Plasma membrane|Secreted", loc)) score <- score + 2
  if (!is.na(hr_val) && hr_val > 1.5) score <- score + 3
  if (!is.na(logFC_disc) && logFC_disc > 2.0) score <- score + 1
  if (status == "Novel Candidate") score <- score + 2
  
  target_info_list[[gene]] <- data.frame(
    gene_symbol = gene,
    subcellular_localization = loc,
    novelty_status = status,
    modality_category = modality,
    logFC_discovery = logFC_disc,
    logFC_validation = logFC_valid,
    survival_HR = hr_val,
    survival_p_value = p_val,
    priority_score = score,
    stringsAsFactors = FALSE
  )
}

# Bind into dataframe
all_targets_df <- do.call(rbind, target_info_list) %>%
  arrange(desc(priority_score), desc(logFC_discovery))

# Filter for Novel Candidates
potential_novel_targets <- all_targets_df %>%
  filter(novelty_status == "Novel Candidate")

cat("Total novel candidate genes identified:", nrow(potential_novel_targets), "\n")
cat("Displaying prioritized novel target candidates:\n")
print(head(potential_novel_targets, 15))

# Save the full target table and the filtered novel target table
write.csv(all_targets_df, "data_clean/target_druggability_analysis.csv", row.names = FALSE)
write.csv(potential_novel_targets, "data_clean/potential_novel_targets_list.csv", row.names = FALSE)
write.csv(potential_novel_targets, "Week5/potential_novel_targets_list.csv", row.names = FALSE)

cat("\nOutputs successfully saved to data_clean/ and Week5/:\n")
cat("  - Week5/potential_novel_targets_list.csv\n")
cat("  - go/kegg plots in Week5/\n")

cat("\nPipeline script 08 completed successfully!\n")

```

## Section 9: Cross-Dataset Validation in Lymphoid Malignancy (DLBCL)

Downloads GSE10846, evaluates the prognostic value of candidates in DLBCL, and classifies them into pan-hematological vs. AML-specific targets.

```{r}
# ==============================================================================
# SECTION 0: Setup and Package Loading | 第 0 部分：准备与加载包
# ==============================================================================

if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

required_bioc <- c("GEOquery", "Biobase")
required_cran <- c("data.table", "dplyr", "stringr", "survival", "ggplot2", "gridExtra")

for (pkg in required_bioc) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    BiocManager::install(pkg)
  }
}

for (pkg in required_cran) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    install.packages(pkg)
  }
}

library(GEOquery)
library(Biobase)
library(data.table)
library(dplyr)
library(stringr)
library(survival)
library(ggplot2)
library(gridExtra)

# Create output directories
dir.create("Week6", showWarnings = FALSE)
dir.create("results", showWarnings = FALSE)

# ==============================================================================
# SECTION 1: Load and Preprocess GSE10846 Dataset | 第 1 部分：加载并预处理 GSE10846
# ==============================================================================

eset_file <- "data_raw/gse10846/gse10846_eset.rds"
if (!file.exists(eset_file)) {
  dir.create("data_raw/gse10846", recursive = TRUE, showWarnings = FALSE)
  options(timeout = 100000)
  cat("Downloading GSE10846 Series Matrix from GEO...\n")
  gse_list <- getGEO(GEO = "GSE10846", GSEMatrix = TRUE, getGPL = FALSE, destdir = "data_raw/gse10846")
  if ("GPL570" %in% names(gse_list)) {
    gse10846_eset <- gse_list[["GPL570"]]
  } else {
    gse10846_eset <- gse_list[[1]]
  }
  saveRDS(gse10846_eset, file = eset_file)
} else {
  cat("Loading local GSE10846 ExpressionSet...\n")
  gse10846_eset <- readRDS(eset_file)
}

gse_expr_raw <- exprs(gse10846_eset)
gse_pheno <- pData(gse10846_eset)

# Map Probes to Gene Symbols
gpl_file <- "data_raw/gse13159/GPL570.annot.gz"
if (!file.exists(gpl_file)) {
  dir.create(dirname(gpl_file), recursive = TRUE, showWarnings = FALSE)
  download.file(
    url = "https://ftp.ncbi.nlm.nih.gov/geo/platforms/GPLnnn/GPL570/annot/GPL570.annot.gz",
    destfile = gpl_file,
    mode = "wb"
  )
}

gpl_data <- parseGEO(gpl_file)
gpl_table <- Table(gpl_data)

probe_gene_info <- data.frame(
  probe_id = as.character(gpl_table$ID),
  gene_symbol_raw = as.character(gpl_table$`Gene symbol`),
  stringsAsFactors = FALSE
) %>%
  mutate(
    gene_symbol = str_split(gene_symbol_raw, "///", simplify = TRUE)[, 1],
    gene_symbol = str_trim(gene_symbol)
  ) %>%
  filter(!is.na(gene_symbol), gene_symbol != "", gene_symbol != "---", gene_symbol != "NA")

expr_dt <- as.data.table(gse_expr_raw, keep.rownames = "probe_id")
probe_gene_dt <- as.data.table(probe_gene_info)
expr_gene_dt <- merge(expr_dt, probe_gene_dt[, .(probe_id, gene_symbol)], by = "probe_id")

sample_cols <- colnames(gse_expr_raw)
gse_expr_by_gene <- expr_gene_dt[, lapply(.SD, mean, na.rm = TRUE), by = gene_symbol, .SDcols = sample_cols]

gse_expr_clean <- as.data.frame(gse_expr_by_gene)
rownames(gse_expr_clean) <- gse_expr_clean$gene_symbol
gse_expr_clean$gene_symbol <- NULL

# Truncate negative values and apply log2 transformation if not log scale
expr_range <- range(gse_expr_clean, na.rm = TRUE)
if (expr_range[2] > 100) {
  gse_expr_clean[gse_expr_clean < 0] <- 0
  gse_expr_final <- log2(gse_expr_clean + 1)
} else {
  gse_expr_final <- gse_expr_clean
}

saveRDS(gse_expr_final, file = "data_clean/gse10846_expr_clean.rds")

# ==============================================================================
# SECTION 2: Parse and Clean Phenotype Data | 第 2 部分：解析临床生存数据
# ==============================================================================

# Parse characteristics columns dynamically to avoid column offset errors
cat("Parsing DLBCL clinical data...\n")
parsed_clinical <- do.call(rbind, lapply(1:nrow(gse_pheno), function(i) {
  row_text <- paste(gse_pheno[i, ], collapse = "; ")
  status <- str_trim(str_match(row_text, "Follow up status: ([^;]+)")[, 2])
  years <- as.numeric(str_trim(str_match(row_text, "Follow up years: ([^;]+)")[, 2]))
  chemo <- str_trim(str_match(row_text, "Chemotherapy: ([^;]+)")[, 2])
  diag <- str_trim(str_match(row_text, "Final microarray diagnosis: ([^;]+)")[, 2])
  data.frame(
    sample_id = rownames(gse_pheno)[i],
    status = status,
    years = years,
    chemo = chemo,
    diag = diag,
    stringsAsFactors = FALSE
  )
}))

parsed_clinical <- parsed_clinical %>%
  mutate(
    event = ifelse(status == "DEAD", 1, ifelse(status == "ALIVE", 0, NA))
  )

# Filter samples with valid survival data and standard chemotherapy regimens
valid_idx <- which(!is.na(parsed_clinical$event) & !is.na(parsed_clinical$years) & 
                     parsed_clinical$chemo %in% c("CHOP-Like Regimen", "R-CHOP-Like Regimen"))

clean_expr <- gse_expr_final[, parsed_clinical$sample_id[valid_idx]]
clean_cli <- parsed_clinical[valid_idx, ]
clean_cli$chemo <- factor(clean_cli$chemo, levels = c("CHOP-Like Regimen", "R-CHOP-Like Regimen"))

saveRDS(clean_cli, file = "data_clean/gse10846_sample_info.rds")

# ==============================================================================
# SECTION 3: Perform Survival Analysis | 第 3 部分：执行生存分析
# ==============================================================================

targets <- c("IL1R2", "VNN2", "SLC15A3", "EGFL7", "CXCR1", "CXCR2", "CD14", "GZMB", "FPR2", "CMTM2")
dlbcl_survival_results <- list()

cat("Running survival analyses for top target candidates in DLBCL...\n")
for (g in targets) {
  val <- as.numeric(clean_expr[g, ])
  med <- median(val, na.rm = TRUE)
  grp <- ifelse(val >= med, "High", "Low")
  
  # Aligned clinical variables
  df_temp <- clean_cli
  df_temp$gene_expr <- val
  df_temp$gene_group <- factor(grp, levels = c("Low", "High"))
  
  # Log-rank test
  logrank_fit <- survdiff(Surv(years, event) ~ gene_group, data = df_temp)
  df <- length(logrank_fit$n) - 1
  logrank_p <- pchisq(logrank_fit$chisq, df = df, lower.tail = FALSE)
  
  # Univariate Cox
  uni_fit <- coxph(Surv(years, event) ~ gene_group, data = df_temp)
  uni_summary <- summary(uni_fit)
  uni_hr <- uni_summary$conf.int[1, 1]
  uni_lower <- uni_summary$conf.int[1, 3]
  uni_upper <- uni_summary$conf.int[1, 4]
  uni_p <- uni_summary$waldtest[3]
  
  # Multivariate adjusted Cox (adjust for chemo regimen)
  adj_fit <- coxph(Surv(years, event) ~ gene_group + chemo, data = df_temp)
  adj_summary <- summary(adj_fit)
  adj_hr <- adj_summary$conf.int[1, 1]
  adj_lower <- adj_summary$conf.int[1, 3]
  adj_upper <- adj_summary$conf.int[1, 4]
  adj_p <- adj_summary$coefficients[1, 5]
  
  dlbcl_survival_results[[g]] <- data.frame(
    gene_symbol = g,
    logrank_p = logrank_p,
    unadj_HR = uni_hr,
    unadj_HR_lower = uni_lower,
    unadj_HR_upper = uni_upper,
    unadj_cox_p = uni_p,
    adj_HR = adj_hr,
    adj_HR_lower = adj_lower,
    adj_HR_upper = adj_upper,
    adj_cox_p = adj_p,
    stringsAsFactors = FALSE
  )
}

dlbcl_res_df <- do.call(rbind, dlbcl_survival_results)
write.csv(dlbcl_res_df, file = "data_clean/gse10846_dlbcl_survival_results.csv", row.names = FALSE)
write.csv(dlbcl_res_df, file = "Week6/gse10846_dlbcl_survival_results.csv", row.names = FALSE)

print(dlbcl_res_df)

# ==============================================================================
# SECTION 4: Draw Kaplan-Meier Curves for Validated Targets | 第 4 部分：绘制 K-M 生存曲线
# ==============================================================================

# Custom ggplot2 KM plotting function to avoid survminer dependencies
draw_km_ggplot <- function(gene, data, clean_expr_data) {
  val <- as.numeric(clean_expr_data[gene, ])
  med <- median(val, na.rm = TRUE)
  grp <- ifelse(val >= med, "High", "Low")
  
  df_temp <- data
  df_temp$gene_group <- factor(grp, levels = c("Low", "High"))
  
  # Fit survival curve
  fit <- survfit(Surv(years, event) ~ gene_group, data = df_temp)
  
  # Compile plot dataframe
  fit_df <- data.frame(
    time = fit$time,
    surv = fit$surv,
    upper = fit$upper,
    lower = fit$lower,
    n.censor = fit$n.censor,
    group = rep(names(fit$strata), fit$strata)
  )
  fit_df$group <- gsub("gene_group=", "", fit_df$group)
  
  # Extract p-value
  p_val <- dlbcl_res_df[dlbcl_res_df$gene_symbol == gene, "logrank_p"]
  p_text <- paste0("Log-rank P = ", format.pval(p_val, digits = 3))
  
  # Plot
  p <- ggplot(fit_df, aes(x = time, y = surv, color = group)) +
    geom_step(size = 1.2) +
    labs(
      title = paste0("GSE10846 (DLBCL): ", gene),
      x = "Survival Time (Years)",
      y = "Overall Survival Probability",
      color = "Expression Group"
    ) +
    scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
    scale_color_manual(values = c("High" = "#D95F02", "Low" = "#1B9E77")) +
    annotate("text", x = 1.5, y = 0.15, label = p_text, fontface = "italic", size = 4.5, color = "black") +
    theme_classic() +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold", size = 13),
      axis.title = element_text(size = 11),
      legend.position = "bottom",
      panel.grid.major.y = element_line(color = "gray90", linetype = "dashed")
    )
  return(p)
}

cat("Generating Kaplan-Meier plots for validated targets...\n")
p_vnn2 <- draw_km_ggplot("VNN2", clean_cli, clean_expr)
p_fpr2 <- draw_km_ggplot("FPR2", clean_cli, clean_expr)
p_cmtm2 <- draw_km_ggplot("CMTM2", clean_cli, clean_expr)

# Arrange and save as a panel
g_km <- grid.arrange(p_vnn2, p_fpr2, p_cmtm2, ncol = 3)
ggsave(filename = "Week6/km_curves_dlbcl_validated.png", plot = g_km, width = 12, height = 4.2, dpi = 300)
ggsave(filename = "results/km_curves_dlbcl_validated.png", plot = g_km, width = 12, height = 4.2, dpi = 300)

# ==============================================================================
# SECTION 5: Draw Cross-Dataset HR Forest Plot | 第 5 部分：绘制跨数据集 HR 森林对比图
# ==============================================================================

# Load AML survival results
aml_res <- read.csv("data_clean/survival_cox_results.csv") %>%
  filter(gene_symbol %in% targets) %>%
  select(gene_symbol, HR, HR_CI_lower, HR_CI_upper, p_value) %>%
  mutate(Cohort = "AML (TCGA LAML)")

# Clean columns and bind
dlbcl_forest_data <- dlbcl_res_df %>%
  select(gene_symbol, adj_HR, adj_HR_lower, adj_HR_upper, adj_cox_p) %>%
  rename(HR = adj_HR, HR_CI_lower = adj_HR_lower, HR_CI_upper = adj_HR_upper, p_value = adj_cox_p) %>%
  mutate(Cohort = "DLBCL (GSE10846, Adjusted)")

forest_all <- rbind(aml_res, dlbcl_forest_data)

# Reorder targets logically (e.g. consistently significant, AML-specific, lineage-opposing)
forest_all$gene_symbol <- factor(forest_all$gene_symbol, levels = rev(c("CMTM2", "FPR2", "GZMB", "IL1R2", "SLC15A3", "CXCR1", "CXCR2", "CD14", "EGFL7", "VNN2")))

p_forest <- ggplot(forest_all, aes(x = HR, y = gene_symbol, color = Cohort)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "gray50") +
  geom_errorbarh(aes(xmin = HR_CI_lower, xmax = HR_CI_upper), height = 0.3, size = 0.8, position = position_dodge(0.5)) +
  geom_point(size = 3.5, position = position_dodge(0.5)) +
  labs(
    title = "Cross-Dataset Prognostic Hazard Comparison (AML vs. DLBCL)",
    x = "Hazard Ratio (HR) with 95% CI",
    y = "Target Candidates",
    color = "Malignancy Cohort"
  ) +
  scale_color_manual(values = c("AML (TCGA LAML)" = "#E41A1C", "DLBCL (GSE10846, Adjusted)" = "#377EB8")) +
  theme_classic() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 12),
    axis.title = element_text(size = 10),
    axis.text = element_text(size = 10),
    legend.position = "right"
  )

ggsave(filename = "Week6/forest_plot_cross_dataset.png", plot = p_forest, width = 8.5, height = 5.2, dpi = 300)
ggsave(filename = "results/forest_plot_cross_dataset.png", plot = p_forest, width = 8.5, height = 5.2, dpi = 300)

cat("All validations and visualizations completed successfully!\n")

```

